# The Model {#sec-model}

```{r}
#| label: setup-02
#| include: false
library(countSFA)
library(ggplot2)
library(patchwork)
```

Let $Y_i$ denote a non-negative integer output for unit $i = 1, \ldots, n$, and let $\mathbf{x}_i \in \mathbb{R}^k$ be a vector of observed inputs. The frontier specification posits a maximum (or minimum) expected output level $\lambda_i = \lambda(\mathbf{x}_i;\, \boldsymbol{\beta}) > 0$ from which each unit's realised output is displaced by a one-sided random factor. We treat production and cost frontiers in parallel through

$$
Y_i \mid u_i,\, \mathbf{x}_i \;\sim\;
   \mathrm{Poisson}\!\left(\lambda(\mathbf{x}_i;\boldsymbol{\beta})\, e^{\mp u_i}\right),
\qquad u_i \geq 0,
$$ {#eq-conditional}

where the upper sign defines the *production frontier*, where $\lambda_i$ is an upper bound on the conditional mean, and the lower sign defines the *cost frontier*, where $\lambda_i$ is a lower bound. Throughout, $\boldsymbol{\beta}$ is the frontier parameter vector, $u_i \geq 0$ is the technical inefficiency, and the triples $(Y_i, u_i, \mathbf{x}_i)$ are independent across units. We write

$$
a_i \;\equiv\; a(\mathbf{x}_i;\,\boldsymbol{\beta})
   \;=\; \log\lambda(\mathbf{x}_i;\,\boldsymbol{\beta})
$$ {#eq-loglink}

for the log-frontier mean. The closed-form derivations that follow require only that $\lambda(\mathbf{x}_i;\,\boldsymbol{\beta})$ be positive, and they make no use of any specific functional form. For estimation, simulation, and the empirical application we adopt the canonical log-linear specification $\lambda(\mathbf{x}_i;\,\boldsymbol{\beta}) = \exp(\mathbf{x}_i'\boldsymbol{\beta})$, under which $a_i = \mathbf{x}_i'\boldsymbol{\beta}$. Nonlinear or semiparametric forms are admissible at no cost to the marginal PMF or to the conditional efficiency score.

The specification follows the count-data frontier of @feHofler2013. In either orientation, however, their half-normal inefficiency leaves the marginal PMF as a one-dimensional integral that the package evaluates by maximum simulated likelihood with Halton sequences. The Gamma family adopted below yields a closed-form alternating series in the production case and a one-dimensional integral with a uniformly damped integrand in the cost case, permitting exact likelihood evaluation without simulation.

## Inefficiency Distribution

Following @greene1990 in the continuous frontier literature, we model

$$
u_i \mid \alpha,\, b \;\sim\; \mathrm{Gamma}(\alpha,\, b),
\qquad \alpha > 0,\; b > 0,
$$ {#eq-gamma-dist}

with density $f(u;\, \alpha, b) = b^\alpha u^{\alpha - 1} e^{-bu} / \Gamma(\alpha)$ for $u \geq 0$ (rate parameterisation) and moments

$$
\mathrm{E}[u_i] \;=\; \frac{\alpha}{b},
\qquad
\mathrm{Var}[u_i] \;=\; \frac{\alpha}{b^2}.
$$ {#eq-gamma-moments}

The shape parameter $\alpha$ governs the *form* of the inefficiency distribution independently of its scale. When $\alpha < 1$ the density is decreasing on $[0, \infty)$ with mode at zero. When $\alpha = 1$ it is exactly exponential, and we recover a count-data analogue of that classical specification of @meeusenBroeck1977. When $\alpha > 1$ the density is hump-shaped with an interior mode at $(\alpha - 1)/b > 0$, implying that very small inefficiency is no more likely than some positive reference level. The latter case is economically important: in sectors where all producers face an irreducible level of inefficiency, e.g., regulatory compliance costs, adjustment lags, or organisational slack, a distribution with an interior mode more accurately reflects the data-generating process. The rate $b$ is a scale parameter: large $b$ concentrates inefficiency near zero so that nearly all units approach the frontier; small $b$ implies widespread and severe inefficiency.

## Marginal PMF

::: {#prp-pmf}

## Marginal PMF under Gamma inefficiency

Let $u_i \sim \mathrm{Gamma}(\alpha, b)$ as in @eq-gamma-dist. Under @eq-conditional, the marginal probability mass function of $Y_i$ takes the following two forms.

*(i) Production frontier* $\bigl(Y_i \mid u_i \sim \mathrm{Poisson}(\lambda_i e^{-u_i})\bigr)$. Marginalising over $u_i$ yields the absolutely convergent alternating-series representation

$$
\Pr(Y_i = y \mid \mathbf{x}_i,\, \alpha,\, b)
\;=\; \frac{b^\alpha\, e^{a_i y}}{y!}
  \sum_{j=0}^{\infty}
    \frac{(-1)^j\, e^{j a_i}}{j!\;(y + b + j)^\alpha},
$$ {#eq-pmf-gamma}

for all $a_i < \infty$, $b > 0$, $\alpha > 0$, and
$y \in \{0, 1, 2, \ldots\}$.

*(ii) Cost frontier* $\bigl(Y_i \mid u_i \sim \mathrm{Poisson}(\lambda_i e^{+u_i})\bigr)$. No analogous closed-form series exists; the marginal PMF is the one-dimensional integral

$$
\Pr(Y_i = y \mid \mathbf{x}_i,\, \alpha,\, b)
\;=\; \frac{b^\alpha\, e^{a_i y}}{y!\,\Gamma(\alpha)}
  \int_0^\infty u^{\alpha - 1}\,
     \exp\!\bigl((y - b)\,u \,-\, e^{a_i + u}\bigr)\,\mathrm{d}u,
$$ {#eq-pmf-gamma-cost}

whose integrand is super-exponentially damped as $u \to \infty$ and integrable at $u = 0^+$ for every $\alpha > 0$, so the integral is finite for all $b > 0$ and can be approximated by adaptive quadrature.
:::

Although @eq-pmf-gamma converges absolutely for every $\alpha > 0$, the alternating signs in $(-1)^j$ cause catastrophic cancellation once $e^{a_i}$ becomes moderate, and the cancellation pattern depends sensitively on $\alpha$: in the sub-exponential regime $\alpha < 1$ the term magnitudes decay more slowly so that summation loses many digits before the tail vanishes, while at the other extreme large $\alpha$ inflates the partial sums in absolute value relative to the small signed total. Signed log-sum-exp and series-acceleration techniques alleviate the loss but do not eliminate it across the full $(a_i, b, \alpha)$ grid encountered in estimation and Monte Carlo. We therefore do not evaluate the series numerically. Instead, we use the integral representation underlying @eq-pmf-gamma. Marginalising @eq-conditional under the production sign convention $Y_i \mid u_i \sim \mathrm{Poisson}(\lambda_i e^{-u_i})$ gives

$$
\Pr(Y_i = y \mid \mathbf{x}_i,\, \alpha,\, b)
\;=\; \frac{b^\alpha\, e^{a_i y}}{y!\,\Gamma(\alpha)}
  \int_0^\infty u^{\alpha - 1}\,
     \exp\!\bigl(-(y + b)\,u \,-\, e^{a_i - u}\bigr)\,\mathrm{d}u,
$$ {#eq-pmf-gamma-prod}

a one-dimensional integral parallel to the cost-orientation expression @eq-pmf-gamma-cost. The integrand is integrable at $u = 0^+$ for every $\alpha > 0$ and super-exponentially damped as $u \to \infty$, so adaptive quadrature returns the production PMF accurately across the full $(a_i, b, \alpha)$ grid. The implementation in §4 uses this quadrature path for both orientations; the alternating-series form @eq-pmf-gamma is retained only for its analytic role in deriving the exponential corollary below.

### The Exponential Corollary {#sec-pmf-exp}

Setting $\alpha = 1$ collapses the alternating series @eq-pmf-gamma to a closed form involving the lower incomplete gamma function. The Taylor expansion of $\gamma(s, x) = \int_0^x t^{s-1} e^{-t}\,\mathrm{d}t$ about $x = 0$ is
$$
\gamma(s,\, x) \;=\; \sum_{j=0}^\infty \frac{(-1)^j x^{s+j}}{j!\,(s+j)},
$$
so evaluating at $s = y + b$ and $x = e^{a_i}$ and multiplying through by $e^{-b a_i}$ gives
$$
e^{-b a_i}\,\gamma(y+b,\, e^{a_i})
\;=\; e^{a_i y} \sum_{j=0}^\infty \frac{(-1)^j e^{j a_i}}{j!\,(y+b+j)},
$$
so that @eq-pmf-gamma at $\alpha = 1$ becomes a multiple of $\gamma(y+b,\, e^{a_i})$. The cost case admits a parallel closed form via the direct change of variable $t = e^{a_i + u}$ in @eq-pmf-gamma-cost at $\alpha = 1$: this sends $u \in [0, \infty)$ to $t \in [e^{a_i}, \infty)$ and collapses the integrand to $t^{y - b - 1}\, e^{-t}\, e^{-(y-b)a_i}$, whose integral is $e^{-(y-b)a_i}\,\Gamma(y - b,\,e^{a_i})$.

::: {#cor-pmf-exp}

## Exponential closed form

When $\alpha = 1$ the marginal PMFs reduce to
$$
\Pr(Y_i = y \mid \mathbf{x}_i,\, b)
\;=\; \frac{b\, e^{-b a_i}}{y!}\;\gamma\!\left(y + b,\; e^{a_i}\right)
$$ {#eq-pmf-exp}

in the production frontier, and to
$$
\Pr(Y_i = y \mid \mathbf{x}_i,\, b)
\;=\; \frac{b\, e^{a_i b}}{y!}\;\Gamma\!\left(y - b,\; e^{a_i}\right)
$$ {#eq-cost-pmf}

in the cost frontier, where $\gamma(s, x) = \int_0^x t^{s-1} e^{-t}\,\mathrm{d}t$ and $\Gamma(s, x) = \int_x^\infty t^{s-1} e^{-t}\,\mathrm{d}t$ denote the lower and upper incomplete gamma functions respectively.
:::

The two cases are structurally symmetric: the sign of $a_i b$ in the multiplying factor flips, the lower and upper incomplete gamma exchange, and to our knowledge the cost-frontier closed form @eq-cost-pmf has not previously been derived for count-data frontiers. In the half-normal model of @feHofler2013 both orientations require simulated likelihood, whereas here both admit exact evaluation through standard special functions.

::: {.remark title="Frontier constraint and the incomplete gamma"}
The incompleteness of the gamma function in @eq-pmf-exp and @eq-cost-pmf is not a computational device but the algebraic consequence of the one-sided constraint $u_i \geq 0$. In the production case the change of variable $s = \lambda_i e^{-u_i}$ runs over $(0, \lambda_i]$, restricting the resulting integral to the lower incomplete gamma; in the cost case $t = \lambda_i e^{+u_i}$ runs over $[\lambda_i, \infty)$ and produces the upper incomplete gamma. Relaxing $u_i \geq 0$ to $u_i \in \mathbb{R}$ would extend each integral to all of $(0, \infty)$, replacing the incomplete gamma with the complete $\Gamma(y+b)$, and recovering the negative binomial PMF of an unconstrained Poisson--Gamma mixture [@cameronTrivedi2005, Ch. 20]. The frontier model is therefore econometrically distinct from a negative binomial regression precisely because the gamma function in its likelihood is incomplete.
:::

::: {.remark title="Negative shape arguments in the cost-frontier PMF"}
The upper incomplete gamma $\Gamma(s, x) = \int_x^\infty t^{s-1} e^{-t}\,\mathrm{d}t$ is well-defined for any real $s$ whenever $x > 0$: the exponential decay of $e^{-t}$ dominates the polynomial factor $t^{s-1}$ irrespective of the sign of
$s - 1$, and the lower limit $x = e^{a_i} > 0$ keeps the integrand away from any singularity at the origin. For count data with $y \geq 0$ and $b > 0$ the argument $y - b$ in @eq-cost-pmf can be negative (e.g.\ $y = 0$, $b = 1$
gives $s = -1$), but $\Gamma(y - b,\, e^{a_i})$ remains finite. Standard regularised-incomplete-gamma routines distributed with most statistical software require shape $> 0$; the unnormalised upper incomplete gamma at arbitrary real shape is available through GSL-compatible numerical libraries via the standard recurrence and continued-fraction expansions, or one can evaluate the integral @eq-pmf-gamma-cost directly by quadrature.
:::

@fig-pmf-shape illustrates the marginal PMF under two parameter sweeps at fixed frontier mean $\lambda_i = \exp(a_i) = 8$. The left panel shows the exponential case ($\alpha = 1$) as the rate $b$ decreases from 5 toward 0.5: probability mass shifts leftward and the distribution becomes progressively heavier-tailed compared with the Poisson reference, reflecting the increasing prevalence and severity of inefficiency. The right panel shows how the shape parameter $\alpha$ at fixed $b = 1$ reshapes the same distribution: as $\alpha$ rises from $0.5$ (decreasing inefficiency density) toward $4$ (strongly hump-shaped density), the PMF concentrates because the inefficiency density rules out near-zero values; conversely $\alpha < 1$ produces a
heavier left tail than the exponential, as the monotone-decreasing density assigns positive probability to arbitrarily large inefficiency values.

```{r}
#| label: fig-pmf-shape
#| fig-width: 8
#| fig-height: 3.5
#| out-width: "100%"
#| fig-cap: "Marginal PMF of $Y_i$ for $\\lambda_i = \\exp(a_i) = 8$ under two parameter sweeps. Left panel: exponential inefficiency ($\\alpha = 1$), four rate values $b \\in \\{0.5, 1, 2, 5\\}$. Right panel: Gamma inefficiency at fixed $b = 1$, four shape values $\\alpha \\in \\{0.5, 1, 2, 4\\}$. Grey bars: Poisson($\\lambda_i$) reference, attained as $b \\to \\infty$ (left panel) or as $\\alpha \\to \\infty$ at fixed mean (right panel)."

a_val      <- log(8)
y_seq      <- 0:25
b_vals     <- c(0.5, 1, 2, 5)
alpha_vals <- c(0.5, 1, 2, 4)

# Exponential sweep: vary b at alpha = 1
df_b <- do.call(rbind, lapply(b_vals, function(b_val) {
  pmf_vals <- sapply(y_seq, function(y)
    exp(pmf_poisson_gamma(y, a_val, b = b_val, alpha = 1L)))
  data.frame(y = y_seq, pmf = pmf_vals,
             lvl = factor(sprintf("b == %g", b_val),
                          levels = sprintf("b == %g", b_vals)))
}))

# Gamma sweep: vary alpha at b = 1
df_a <- do.call(rbind, lapply(alpha_vals, function(al) {
  pmf_vals <- sapply(y_seq, function(y)
    exp(pmf_poisson_gamma(y, a_val, b = 1, alpha = al)))
  data.frame(y = y_seq, pmf = pmf_vals,
             lvl = factor(sprintf("alpha == %g", al),
                          levels = sprintf("alpha == %g", alpha_vals)))
}))

# Shared Poisson reference
pois_ref <- data.frame(y = y_seq, pmf = dpois(y_seq, lambda = exp(a_val)))

# Panel builder: one PMF overlay on the Poisson reference, with the
# parameter sweep encoded by colour, linetype, and shape; legend placed
# inside the panel so each subplot is self-contained.
mk_panel <- function(df, title) {
  ggplot() +
    geom_col(data = pois_ref, aes(x = y, y = pmf),
             fill = "grey80", colour = "grey60", width = 0.8) +
    geom_line(data = df,
              aes(x = y, y = pmf, colour = lvl, linetype = lvl),
              linewidth = 0.7) +
    geom_point(data = df,
               aes(x = y, y = pmf, colour = lvl, shape = lvl),
               size = 1.4) +
    scale_colour_brewer(palette = "Dark2", name = NULL,
                        labels = scales::parse_format()) +
    scale_linetype_discrete(name = NULL,
                            labels = scales::parse_format()) +
    scale_shape_discrete(name = NULL,
                         labels = scales::parse_format()) +
    labs(title = title,
         x = expression(italic(y)), y = "Probability") +
    theme_bw(base_size = 10) +
    theme(legend.position = "inside",
          legend.position.inside = c(0.82, 0.72),
          legend.background = element_rect(fill = "white", colour = NA),
          plot.title = element_text(size = 10))
}

p_b <- mk_panel(df_b, "Exponential (\u03b1 = 1): varying b")
p_a <- mk_panel(df_a, "Gamma (b = 1): varying \u03b1")

p_b + p_a + plot_layout(ncol = 2)
```

## Technical Efficiency Scores

Following standard practice in stochastic frontier analysis [@kumbhakarLovell2000, Ch. 2], the technical efficiency of unit $i$ is defined as the conditional expectation of the efficiency factor $e^{-u_i}$ given the observed count,

$$
\widehat{\mathrm{TE}}_i
\;\equiv\; \mathrm{E}\!\left[e^{-u_i} \mid Y_i = y_i,\; \mathbf{x}_i\right].
$$ {#eq-te-def}

This decomposition separates the one-sided inefficiency from the Poisson noise using the full likelihood structure [@JLMS:1982].

::: {#prp-te}

## Conditional efficiency score under Gamma inefficiency

Under the Gamma inefficiency assumption @eq-gamma-dist and the conditional model @eq-conditional, the conditional technical efficiency takes the following two forms.

*(i) Production frontier.*

$$
\mathrm{E}\!\left[e^{-u_i} \mid Y_i = y_i,\, \mathbf{x}_i\right]
\;=\;
\frac{\displaystyle\sum_{j=0}^\infty
        \dfrac{(-1)^j e^{j a_i}}{j!\,(y_i + b + 1 + j)^\alpha}}
     {\displaystyle\sum_{j=0}^\infty
        \dfrac{(-1)^j e^{j a_i}}{j!\,(y_i + b + j)^\alpha}}.
$$ {#eq-te-gamma}

*(ii) Cost frontier.*

$$
\mathrm{E}\!\left[e^{-u_i} \mid Y_i = y_i,\, \mathbf{x}_i\right]
\;=\;
\frac{\displaystyle\int_0^\infty u^{\alpha-1}\,
        \exp\!\bigl((y_i - b - 1)\,u - e^{a_i + u}\bigr)\,\mathrm{d}u}
     {\displaystyle\int_0^\infty u^{\alpha-1}\,
        \exp\!\bigl((y_i - b)\,u - e^{a_i + u}\bigr)\,\mathrm{d}u}.
$$ {#eq-te-gamma-cost}

Both integrals are finite for all $b > 0$ and $\alpha > 0$ by the finiteness argument used for @eq-pmf-gamma-cost; numerical quadrature is required for the cost case.
:::

The score lies in $(0, 1)$ for all parameter values: $u_i \geq 0$ implies $e^{-u_i} \in (0, 1]$ almost surely, so the conditional mean is too. The dependence on the data enters entirely through $y_i$ (the observed count) and $a_i$ (the fitted
log-frontier for unit $i$): a unit with $y_i = 0$ but a large frontier mean $\lambda_i$ receives a low score in the production case, correctly diagnosing severe inefficiency; as $b \to \infty$, the score converges uniformly to $1$ in either orientation.

### The Exponential Corollary {#sec-te-exp}

At $\alpha = 1$ the production series in @eq-te-gamma satisfies $\sum_{j=0}^\infty (-1)^j e^{ja_i}/[j!(y_i+b+1+j)] = e^{-(y_i+b+1)a_i}\,\gamma(y_i+b+1,\, e^{a_i})$ by the same Taylor identity used in @cor-pmf-exp, and analogously the denominator equals $e^{-(y_i+b)a_i}\,\gamma(y_i+b,\, e^{a_i})$. The cost-frontier integrals admit the parallel closed form via the direct substitution $t = e^{a_i + u}$, which yields $\int_0^\infty \exp\bigl((y_i - b - 1)u - e^{a_i + u}\bigr)\,\mathrm{d}u = e^{-(y_i - b - 1)a_i}\,\Gamma(y_i - b - 1,\, e^{a_i})$ in the numerator and $e^{-(y_i - b)a_i}\,\Gamma(y_i - b,\, e^{a_i})$ in the denominator.

::: {#cor-te-exp}

## Exponential closed form

When $\alpha = 1$ the conditional efficiency score reduces to
$$
\mathrm{E}\!\left[e^{-u_i} \mid Y_i = y_i,\, \mathbf{x}_i\right]
\;=\; e^{-a_i}\cdot
  \frac{\gamma\!\left(y_i + b + 1,\; e^{a_i}\right)}
       {\gamma\!\left(y_i + b,\; e^{a_i}\right)}
$$ {#eq-te-formula}

in the production frontier, and to
$$
\mathrm{E}\!\left[e^{-u_i} \mid Y_i = y_i,\, \mathbf{x}_i\right]
\;=\; e^{a_i}\cdot
  \frac{\Gamma\!\left(y_i - b - 1,\; e^{a_i}\right)}
       {\Gamma\!\left(y_i - b,\; e^{a_i}\right)}
$$ {#eq-te-formula-cost}

in the cost frontier.
:::

The structural symmetry of @cor-pmf-exp carries over: the sign on $a_i$ in the prefactor flips, and the lower and upper incomplete gamma exchange (with the shape arguments $y_i + b$, $y_i + b + 1$ replaced by $y_i - b$, $y_i - b - 1$). Both forms require no quadrature.

@fig-te-shape illustrates how the conditional efficiency score $\widehat{\mathrm{TE}}_i$ depends on the observed count $y_i$ at fixed log-frontier $a_i = \log 8$. The left panel fixes $\alpha = 1$ (exponential case) and varies $b$: higher $b$ (less inefficiency on average) compresses the score toward $1$ uniformly. The right panel fixes $b = 1$ (Gamma case) and varies $\alpha$: when $\alpha > 1$, the hump-shaped inefficiency density implies that units with low $y_i$ need not be viewed as severely inefficient, so the score is uniformly higher at low counts than under the exponential.

```{r}
#| label: fig-te-shape
#| fig-width: 8
#| fig-height: 3.5
#| out-width: "100%"
#| fig-cap: "Conditional technical efficiency score $\\widehat{\\mathrm{TE}}_i = \\mathrm{E}[e^{-u_i} \\mid Y_i = y_i,\\, \\mathbf{x}_i]$ as a function of observed count $y_i$, for fixed $a_i = \\log 8$. Left panel: exponential inefficiency ($\\alpha = 1$), four rate values $b \\in \\{0.5, 1, 2, 5\\}$; higher $b$ compresses the score toward $1$. Right panel: Gamma inefficiency at fixed $b = 1$, four shape values $\\alpha \\in \\{0.5, 1, 2, 4\\}$; $\\alpha > 1$ raises the score at low counts because the hump-shaped inefficiency density makes some positive $u_i$ expected, so observing $y_i = 0$ is less diagnostic of severe inefficiency. Dotted line: $\\widehat{\\mathrm{TE}} = 1$."

# Left panel: exponential case (alpha = 1), closed-form via lgamma + pgamma.
te_b <- do.call(rbind, lapply(b_vals, function(b_val) {
  te_vals <- sapply(y_seq, function(y) {
    log_r <- lgamma(y + b_val + 1) - lgamma(y + b_val) +
      pgamma(exp(a_val), shape = y + b_val + 1, rate = 1, log.p = TRUE) -
      pgamma(exp(a_val), shape = y + b_val,     rate = 1, log.p = TRUE)
    exp(-a_val + log_r)
  })
  data.frame(y = y_seq, te = te_vals,
             lvl = factor(sprintf("b == %g", b_val),
                          levels = sprintf("b == %g", b_vals)))
}))

# Right panel: Gamma case at b = 1, alternating-series evaluator with
# log-sum-exp guards.
te_series <- function(y, a, b, alpha, delta = 0, K = 150L) {
  kk <- 0:K
  log_abs <- kk * a - lgamma(kk + 1) - alpha * log(y + b + delta + kk)
  pos <- kk %% 2L == 0L
  lse <- function(lv) { m <- max(lv); m + log(sum(exp(lv - m))) }
  lse_p <- lse(log_abs[pos])
  lse_n <- lse(log_abs[!pos])
  if (lse_p <= lse_n) return(NA_real_)
  exp(lse_p + log1p(-exp(lse_n - lse_p)))
}

te_a <- do.call(rbind, lapply(alpha_vals, function(al) {
  te_vals <- sapply(y_seq, function(y) {
    num <- te_series(y, a_val, b = 1, alpha = al, delta = 1)
    den <- te_series(y, a_val, b = 1, alpha = al, delta = 0)
    if (is.na(num) || is.na(den) || den <= 0) return(NA_real_)
    num / den
  })
  data.frame(y = y_seq, te = te_vals,
             lvl = factor(sprintf("alpha == %g", al),
                          levels = sprintf("alpha == %g", alpha_vals)))
}))

# Panel builder: efficiency-score curves with the dotted TE = 1 reference;
# legend placed inside each panel so the composite is self-contained.
mk_te_panel <- function(df, title) {
  ggplot(df, aes(x = y, y = te, colour = lvl, linetype = lvl)) +
    geom_hline(yintercept = 1, linetype = "dotted", colour = "grey50") +
    geom_line(linewidth = 0.85, na.rm = TRUE) +
    geom_point(aes(shape = lvl), size = 1.4, na.rm = TRUE) +
    scale_colour_brewer(palette = "Dark2", name = NULL,
                        labels = scales::parse_format()) +
    scale_linetype_discrete(name = NULL,
                            labels = scales::parse_format()) +
    scale_shape_discrete(name = NULL,
                         labels = scales::parse_format()) +
    scale_y_continuous(limits = c(0, 1.02), breaks = seq(0, 1, by = 0.2)) +
    labs(title = title,
         x = expression(Observed~count~italic(y[i])),
         y = expression(widehat(TE)[i])) +
    theme_bw(base_size = 10) +
    theme(legend.position = "inside",
          legend.position.inside = c(0.82, 0.28),
          legend.background = element_rect(fill = "white", colour = NA),
          plot.title = element_text(size = 10))
}

p_te_b <- mk_te_panel(te_b, "Exponential (\u03b1 = 1): varying b")
p_te_a <- mk_te_panel(te_a, "Gamma (b = 1): varying \u03b1")

p_te_b + p_te_a + plot_layout(ncol = 2)
```
