# Maximum Likelihood Estimation {#sec-estimation}

```{r}
#| label: setup-04
#| include: false
library(countSFA)
library(ggplot2)
```

## Log-Likelihood

Let $\boldsymbol{\theta} = (\boldsymbol{\beta}', b, \alpha)'$ collect all free parameters. Given an IID sample $(y_1, \mathbf{x}_1), \ldots, (y_n, \mathbf{x}_n)$ the log-likelihood is equal to

$$
\ell(\boldsymbol{\theta})
\;=\; \sum_{i=1}^n \log \Pr(Y_i = y_i \mid \mathbf{x}_i,\, \boldsymbol{\theta}),
$$ {#eq-loglik}

where the summands are evaluated via @eq-pmf-exp (exponential case, $\alpha = 1$) or @eq-pmf-gamma (general Gamma). Both PMFs are strictly positive and log-concave in $\boldsymbol{\beta}$ for fixed $(b, \alpha)$, which makes the
profile over $\boldsymbol{\beta}$ well-behaved. The joint surface over all parameters is not globally concave, however, so initialisation and reparametrisation are required.

The parameter space is $\boldsymbol{\beta} \in \mathbb{R}^k$, $b > 0$, $\alpha > 0$. To impose positivity of $b$ and $\alpha$ without constrained optimisation, we optimise over the unconstrained reparametrisation

$$
\boldsymbol{\vartheta}
\;=\; \bigl(\boldsymbol{\beta}',\; \log b,\; \log \alpha\bigr)',
$$ {#eq-reparam}

with $\boldsymbol{\vartheta} \in \mathbb{R}^{k+2}$ for the Gamma model and $\boldsymbol{\vartheta} \in \mathbb{R}^{k+1}$ for the exponential ($\alpha$ fixed at 1). The objective passed to the numerical optimiser is thus the negative
log-likelihood $-\ell(\boldsymbol{\vartheta})$, to be minimised.

::: {.remark title="Series truncation (production orientation)"}
The PMF in @eq-pmf-gamma is evaluated by truncating the alternating series at $K = 1,000$ terms. The ratio-test bound from @sec-model implies that $K \approx 3 e^{a_i}$ terms suffice for the absolute truncation error to
fall below machine epsilon at typical frontier means; the implementation issues a warning if relative convergence is not achieved.
:::

::: {.remark title="Cost-frontier likelihood evaluation"}
The cost-orientation PMF @eq-pmf-gamma-cost does not admit a closed-form alternating series (@prp-pmf, part ii). For $\alpha \neq 1$, or for $\alpha = 1$ with $y_i \leq b$, each per-observation log-PMF is evaluated by adaptive quadrature on the change of variable $v = e^{a_i + u}$, which sends @eq-pmf-gamma-cost to a constant times
$$
\int_{e^{a_i}}^{\infty}
   (\log v - a_i)^{\alpha - 1}\, v^{\,y_i - b - 1}\, e^{-v}\,\mathrm{d}v.
$$ {#eq-cost-quad}
At $\alpha = 1$ with $y_i > b$ the closed form @eq-cost-pmf is used directly; for $y_i \leq b$, we revert to numerical integration. Estimation is correspondingly slower in the cost-frontier case.
:::

## Starting Values

Starting values are essential because the likelihood surface has a degenerate ridge and the dimension of the parameter space is $k + 1$ or $k + 2$. We exploit the fact that the Poisson GLM, which ignores
inefficiency, provides consistent estimates of $\boldsymbol{\beta}$ only up to an additive shift in the intercept, but yields consistent estimates of the slope coefficients. Formally, let $\hat{\boldsymbol{\beta}}_{\text{Pois}}$
denote the standard Poisson MLE. Under the homogeneous DGP, $u_i \perp \mathbf{x}_i$, $\log\mathrm{E}[Y_i \mid \mathbf{x}_i] = a_i + \log\mathrm{E}[e^{\mp u_i}]$, so the slope coefficients are asymptotically unaffected by ignoring $u_i$ while the intercept is biased by $\alpha\,\log\bigl(b/(b+1)\bigr) < 0$ in the production orientation and by $\alpha\,\log\bigl(b/(b-1)\bigr) > 0$ in the cost orientation (the latter requires $b_0 > 1$ for the marginal mean to exist). The bias direction reverses with the orientation but its absolute magnitude is the same log-mean-inefficiency factor.

The starting value for $b$ is obtained from the excess Poisson dispersion. At $\alpha = 1$, @eq-moments gives
$\mathrm{Var}[Y_i] - \mathrm{E}[Y_i] = \lambda_i^2 \cdot b/[(b+1)^2(b+2)]$. We use the initialisation

$$
b_0 \;=\; \frac{\overline{\hat\mu}^2}{\max\bigl(\mathrm{Var}(Y) - \bar{Y},\, \varepsilon\bigr)},
$$ {#eq-starting-b}

where $\hat\mu_i$ are the Poisson fitted values and $\varepsilon > 0$ is a small floor preventing division by near-zero excess variance. This is an heuristic choice: it inverts the simpler relation $\mathrm{Var}[Y] \approx \mathrm{E}[Y] + \mathrm{E}[\mu]^2/b$ rather than the exact moment condition above, and serves only as a starting point. The starting value for $\alpha$ is $\alpha_0 = 1$ (i.e., $\log\alpha_0 = 0$), placing the search at
the exponential sub-model and letting the optimiser find the shape.

The sample-dispersion inversion @eq-starting-b is derived from the production-orientation moments @eq-moments, which exist for all $b > 0$. In the cost orientation, the corresponding marginal moments $\mathrm{E}[Y_i]$ and $\mathrm{Var}[Y_i]$ exist only for $b > 1$ and $b > 2$ respectively (@eq-moments-cost), so the sample mean and variance of $Y_i$ have no finite probability limit when the true $b$ lies in $(0, 2]$ and the moment-based inversion lacks an asymptotic justification.

We therefore fix the cost-orientation starter at $b_0 = e^{0.92} \approx 2.5$, and rely on the warm-start from the exponential fit, together with bound-constrained refinement, to navigate the cost-frontier likelihood. The $\alpha_0 = 1$ exponential starting point is unchanged.

## Numerical Optimisation

**Exponential model.** When $\alpha$ is fixed at 1, the $k + 1$ free parameters are $(\boldsymbol{\beta}', \log b)'$. We maximize the log-likelihood using a gradient descent algorithm with a numerical Hessian computed at the solution.

**Gamma model.** Estimating $\alpha$ freely introduces a qualitatively different optimization challenge. The likelihood surface has a nearly flat ridge along the curve $\alpha \to 0$, $b \to \infty$ while holding $\alpha/b$ constant: as $\alpha \to 0$ the Gamma degenerates toward a point mass at zero, reducing the frontier model to a standard Poisson regression. This ridge does not correspond to a local maximum of practical interest but can trap gradient-based methods if the initial value is poorly chosen.

We address this with three strategies. First, we use a warm start: the exponential model is fitted first, and its solution $(\hat{\boldsymbol{\beta}},\log \hat{b})$ together with $\log \alpha_0 = 0$ provides the starting vector
for the Gamma optimisation. This anchors the search near the exponential sub-model, where the likelihood is concave and the gradient is well-defined.

Second, we restrict $\log b \in [-8, 8]$ (corresponding to $b \in [3 \times 10^{-4},\, 3 \times 10^3]$) and $\log \alpha \in [-2, 4]$ (corresponding to $\alpha \in [0.14,\, 55]$), keeping the optimiser away from the degenerate
boundary. Third, if the optimization algorithm does not converge, a secondary grid of four alternative $\log \alpha$ starting points is tried and the best objective value is retained.

The same numerical approach applies to the cost orientation. The per-evaluation cost is set by the quadrature route described in the cost-frontier likelihood remark above rather than by the alternating-series sum.

::: {.remark title="The degeneracy ridge"}
The ridge $\alpha \to 0$, $b \to \infty$ at constant $\alpha/b$ is not an identification failure of the model but a
near-boundary behaviour of the likelihood. At the true $(\alpha_0, b_0)$ with $\alpha_0 > 0$ finite, the Fisher information matrix is positive definite and the MLE is consistent and asymptotically normal. The issue is purely numerical: for any finite sample the gradient of the log-likelihood may be very shallow near the boundary, making unconstrained optimisation slow or unreliable. The parameter bounds above exclude the problematic region while leaving any interior maximum accessible.
:::

@fig-profile-lik illustrates the profile log-likelihood for $\alpha$ on two simulated DGPs at three sample sizes $n \in \{500, 1000, 2500\}$. For each DGP a single master dataset of size $n_{\max} = 2500$ is generated, and the profile at each smaller $n$ uses the first $n$ observations of that same master dataset; the curves within a panel
therefore differ only in the amount of information available, not in the underlying draws. When the true DGP is exponential ($\alpha_0 = 1$, left panel) the profile is sharply maximised at the truth already at $n = 500$, and the LR-95% interval narrows monotonically as $n$ grows. When the true DGP is Gamma with $\alpha_0 = 2$ and $b_0 = 1$ (right panel) the profile is essentially flat at $n = 500$, becomes resolved near $\alpha_0$ at $n = 1000$, and
yields a tight interval around the truth at $n = 2500$. The weak identification of $\alpha$ is a finite-sample phenomenon under high inefficiency-to-signal ratios, not a structural failure: at $n = 2500$ the Fisher information in the $\alpha$ direction is large enough for the profile to recover the truth in either DGP.

```{r}
#| label: profile-lik
#| cache: true

set.seed(20240315L)
n_max  <- 2500L
n_grid <- c(500L, 1000L, n_max)

# For each DGP, draw one master dataset of size n_max once. Profiles at
# smaller n use the first n observations from the same master dataset, so
# the sharpening of the profile as n grows reflects added information
# rather than different draws.
master_dataset <- function(alpha_true, b_true) {
  z   <- rnorm(n_max)
  u   <- rgamma(n_max, shape = alpha_true, rate = b_true)
  lam <- exp(1 + z - u)
  y   <- rpois(n_max, lam)
  list(z = z, y = y, u = u)
}

# Profile log-likelihood for alpha on the first n_use observations of `data`.
# At each fixed alpha on the grid the inner optimiser profiles out
# (beta, log b) starting from a warm-start chain walked outward from the
# truth, with a cold-start fallback to detect basin drift.
profile_loglik_on <- function(data, n_use, alpha_true) {
  z <- data$z[seq_len(n_use)]
  y <- data$y[seq_len(n_use)]
  X <- cbind(1, z)

  alpha_grid <- seq(0.5, 3, by = 0.25)

  # Method-of-moments warm-start: Poisson-GLM slopes with a bias-corrected
  # intercept, plus (b, alpha) from the alpha-free root equation
  # log R_3(b) / log R_2(b) = log(rhs_3) / log(S_2 + 1). The same
  # routine seeds the package's MLE so the profile curves use a starting
  # point consistent with how a practitioner would fit the model.
  mom_init <- mom_starts(y, X, orientation = "production")
  init_par <- c(mom_init$beta, log(mom_init$b))

  # SE-based bounds on beta around the Poisson-GLM warm-start (matching
  # fit_poisson_frontier()'s default in countSFA >= 0.4.0). Without these,
  # optim can wander into a regime where x_i' beta exceeds the auto-K
  # convergence radius of the alternating series and the inner objective
  # is dominated by truncation noise, dropping the profile point.
  glm_fit  <- suppressWarnings(glm(y ~ X - 1, family = poisson()))
  beta_se  <- summary(glm_fit)$coefficients[, "Std. Error"]
  beta_rad <- pmax(6 * beta_se, 3)
  lower    <- c(coef(glm_fit) - beta_rad, -4)
  upper    <- c(coef(glm_fit) + beta_rad,  4)

  # Auto-K (NULL) lets pmf_poisson_gamma pick the truncation depth per
  # observation; countSFA >= 0.4.0 clamps per-observation log-PMFs to
  # the valid range [-1e6, 0] internally, so the package wrapper
  # returns a smooth finite value even when a few observations have
  # truncation-noise log-PMFs.
  safe_nll <- function(par, y, X, alpha) {
    v <- suppressWarnings(log_lik_poisson_frontier(par, y, X, alpha, K = NULL))
    if (!is.finite(v)) 1e10 else v
  }

  fit_at <- function(al, par_init) {
    par_init <- pmin(pmax(par_init, lower + 1e-6), upper - 1e-6)
    optim(par_init, fn = safe_nll,
          y = y, X = X, alpha = al,
          method  = "L-BFGS-B",
          lower   = lower, upper = upper,
          control = list(maxit = 500, factr = 1e2))
  }

  ord  <- order(abs(alpha_grid - alpha_true))
  ll   <- numeric(length(alpha_grid))
  pars <- vector("list", length(alpha_grid))
  for (k in seq_along(ord)) {
    idx    <- ord[k]
    fitted <- which(!vapply(pars, is.null, logical(1)))
    warm   <- if (length(fitted) == 0L) init_par else
      pars[[fitted[which.min(abs(alpha_grid[fitted] - alpha_grid[idx]))]]]
    cands <- list(
      tryCatch(fit_at(alpha_grid[idx], warm),     error = function(e) NULL),
      tryCatch(fit_at(alpha_grid[idx], init_par), error = function(e) NULL)
    )
    cands <- Filter(function(o) !is.null(o) && is.finite(o$value), cands)
    if (length(cands) == 0L) {
      ll[idx]     <- NA_real_
      pars[[idx]] <- warm
    } else {
      best <- cands[[which.min(vapply(cands, `[[`, numeric(1L), "value"))]]
      ll[idx]     <- if (best$value < 1e9) -best$value else NA_real_
      pars[[idx]] <- best$par
    }
  }
  data.frame(alpha = alpha_grid, loglik = ll,
             dgp   = paste0("alpha", alpha_true),
             n     = n_use)
}

# Build the full (DGP, n) profile grid using nested master datasets.
prof_df <- do.call(rbind, lapply(c(1, 2), function(at) {
  dat <- master_dataset(alpha_true = at, b_true = 1)
  do.call(rbind, lapply(n_grid, function(nn)
    profile_loglik_on(dat, nn, alpha_true = at)))
}))
prof_df <- prof_df[is.finite(prof_df$loglik), ]

# Centre each (DGP, n) curve at its maximum so the LR threshold reads
# directly off the y-axis.
prof_df <- do.call(rbind, lapply(
  split(prof_df, list(prof_df$dgp, prof_df$n), drop = TRUE),
  function(d) {
    d$loglik_c <- d$loglik - max(d$loglik, na.rm = TRUE)
    d
  }))

prof_df$n_lab <- factor(paste0("n = ", format(prof_df$n, big.mark = ",")),
                        levels = paste0("n = ", format(n_grid, big.mark = ",")))

```

```{r}
#| label: fig-profile-lik
#| cache: true
#| fig-cap: "Profile log-likelihood for $\\alpha$ on two simulated DGPs ($b_0 = 1$, $\\boldsymbol{\\beta}_0 = (1, 0.5)'$, $z_i \\sim \\mathcal{N}(0, 1)$), each shown at three sample sizes $n \\in \\{500, 1000, 2500\\}$. Within each panel the curves share the same master dataset (the smaller $n$ uses the first $n$ observations of the $n = 2500$ draw), so differences in curvature reflect added information rather than different random draws. At each fixed $\\alpha$, the coefficients $\\boldsymbol{\\beta}$ and rate $b$ are profiled out by inner MLE. Left panel: true $\\alpha_0 = 1$ (exponential DGP). Right panel: true $\\alpha_0 = 2$ (Gamma DGP). Horizontal dotted line: LR-95% threshold $-\\tfrac{1}{2}\\chi^2_{1,0.95} \\approx -1.92$. Red dashed line: true $\\alpha_0$."


dgp_labels <- c(alpha1 = "alpha[0] == 1",
                alpha2 = "alpha[0] == 2")

vline_df <- data.frame(dgp  = c("alpha1", "alpha2"),
                       xint = c(1, 2))

ggplot(prof_df, aes(x = alpha, y = loglik_c,
                    colour = n_lab, linetype = n_lab, group = n_lab)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 0.9) +
  geom_vline(data = vline_df, aes(xintercept = xint),
             linetype = "dashed", colour = "firebrick", linewidth = 0.6,
             inherit.aes = FALSE) +
  geom_hline(yintercept = -qchisq(0.95, df = 1) / 2,
             linetype = "dotted", colour = "grey40") +
  facet_wrap(~ dgp, labeller = as_labeller(dgp_labels, label_parsed)) +
  scale_colour_manual(values = c("#9ecae1", "#3182bd", "#08306b"),
                      name = NULL) +
  scale_linetype_manual(values = c("dotted", "dashed", "solid"),
                        name = NULL) +
  scale_x_continuous(breaks = seq(0.5, 3, by = 0.5)) +
  coord_cartesian(ylim = c(-6, 0.2)) +
  labs(x = expression(alpha),
       y = expression(log~L(alpha) - log~L(hat(alpha)))) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom")
```

The horizontal dotted line in @fig-profile-lik marks the 95% confidence threshold $-\tfrac{1}{2}\chi^2_{1,0.95} \approx -1.92$: values of $\alpha$ whose profile log-likelihood exceeds this threshold form a likelihood-ratio confidence interval for $\alpha$. Reading the figure along the $n$-axis confirms that weak $\alpha$ identification is a finite-sample phenomenon rather than a structural defect of the model. In the exponential DGP the profile is already sharply peaked at $n = 500$ and tightens further with $n$. In the Gamma DGP, the $n = 500$ profile is flat across nearly the full displayed grid, the $n = 1000$ profile begins to resolve, and by $n = 2500$ the LR-95% interval has narrowed around the truth. The practitioner remark in this section recommends profile-likelihood inference whenever $n$ is in the range where the Wald approximation around $\hat\alpha$ remains unreliable.

## Standard Errors and Inference

Standard errors for all parameters are derived from the numerical Hessian
$\widehat{\mathbf{H}} = -\partial^2 \ell / \partial \boldsymbol{\vartheta}\,
\partial \boldsymbol{\vartheta}'$ evaluated at the MLE. The inverse
$\widehat{\mathbf{V}} = \widehat{\mathbf{H}}^{-1}$ provides the asymptotic
variance of $\hat{\boldsymbol{\vartheta}}$. Since $b$ and $\alpha$ are
estimated on the log scale, standard errors on the natural scale are obtained
via the delta method:

$$
\widehat{\mathrm{SE}}(\hat b)
\;=\; \hat b \cdot \widehat{\mathrm{SE}}(\log \hat b),
\qquad
\widehat{\mathrm{SE}}(\hat\alpha)
\;=\; \hat\alpha \cdot \widehat{\mathrm{SE}}(\log \hat\alpha).
$$ {#eq-delta-se}

The slope coefficients $\boldsymbol{\beta}$ enter linearly, so their standard
errors are read directly from the corresponding diagonal elements of
$\widehat{\mathbf{V}}$.

Under standard regularity conditions (the true parameter vector is in the
interior of the parameter space, the log-likelihood is twice continuously
differentiable, and the information matrix is non-singular), the MLE is
consistent and satisfies

$$
\sqrt{n}\,\bigl(\hat{\boldsymbol{\vartheta}} - \boldsymbol{\vartheta}_0\bigr)
\;\xrightarrow{d}\; \mathcal{N}\!\left(\mathbf{0},\;
  \mathbf{I}(\boldsymbol{\vartheta}_0)^{-1}\right),
$$ {#eq-asymp-normal}

where $\mathbf{I}(\boldsymbol{\vartheta}_0)$ is the Fisher information matrix.
These conditions hold at any interior point of the parameter space; the Monte
Carlo experiment of @sec-simulations confirms that the normal approximation is
accurate for $n \geq 500$ across all DGP configurations considered.

::: {.remark title="Weak identification of $\alpha$"}
In practice, $\alpha$ may be weakly identified: the shape of the inefficiency
distribution is determined by the curvature of the log-likelihood in the
$\alpha$ direction, which is often flat relative to the curvature in the
$(\boldsymbol{\beta}, b)$ directions. The Monte Carlo results in
@sec-simulations document this. Although the bias and RMSE of $\hat\alpha$
shrink at the parametric $\sqrt{n}$ rate, the delta-method 95% CI for $\alpha$
under-covers across the entire grid $n \in \{500, 1000, 2500\}$: the
observed-information Hessian under-states the joint variability of
$(\hat\alpha, \hat b)$ when the $\alpha$-ridge is shallow. We therefore
recommend reporting profile-likelihood intervals for $\alpha$
(@fig-profile-lik) rather than relying solely on the delta-method Wald
interval whenever the shape parameter is of inferential interest.
:::


## Hypothesis Testing

The likelihood ratio (LR) test compares the exponential and Gamma
frontier models. The test statistic

$$
\Lambda
\;=\; 2\bigl[\ell(\hat{\boldsymbol{\theta}}_{\text{Gamma}})
            - \ell(\hat{\boldsymbol{\theta}}_{\text{Exp}})\bigr]
\;\xrightarrow{d}\; \chi^2_1
\quad \text{under } H_0{:}\ \alpha = 1
$$ {#eq-lr-test}

follows a chi-squared distribution with one degree of freedom (the single
restriction $\alpha = 1$), provided the exponential model is estimated
at the same $\boldsymbol{\beta}$ and $b$ values, which is guaranteed because
the exponential is the value of the Gamma log-likelihood evaluated at
$\alpha = 1$ and the jointly optimised $(\boldsymbol{\beta}, b)$.

::: {.remark title="Boundary test"}
The standard regularity condition for the $\chi^2_1$ limit in @eq-lr-test requires that $\alpha_0 = 1$ lies in the *interior* of the parameter space for $\alpha$. Since $\alpha$ is bounded below at zero, $\alpha_0 = 1$ is interior, and the chi-squared reference distribution applies. A more delicate boundary
issue would arise if we tested $H_0: \alpha = 0$ (which would put $\alpha_0$ on the boundary), but we do not consider that degenerate case.
:::

Wald tests for individual parameters follow from @eq-asymp-normal and
@eq-delta-se in the usual way. For the shape parameter specifically, the null
$H_0: \alpha = 1$ can be tested via either the LR statistic @eq-lr-test or
the Wald statistic $(\hat\alpha - 1)^2 / \widehat{\mathrm{Var}}(\hat\alpha)$;
the LR form is generally preferred for its better finite-sample size control
[@cameronTrivedi2005, Ch. 7].

## Computing Efficiency Scores

After estimation, firm-level technical efficiency scores are computed as the
conditional expectation $\widehat{\mathrm{TE}}_i = \mathrm{E}[e^{-u_i} \mid
y_i, \mathbf{x}_i]$ described in @sec-model. The orientation stored on the
fitted object selects the formula. At $\alpha = 1$ the closed
forms are used: @eq-te-formula in the production frontier, and
@eq-te-formula-cost in the cost frontier whenever the shape arguments
$y_i - b$ and $y_i - b - 1$ are both positive (i.e.\ $y_i > b + 1$). For
general $\alpha$, or for $\alpha = 1$ in the cost orientation when
$y_i \leq b + 1$ falls into the negative-shape regime where the
closed-form expression is unavailable, the conditional mean and second
moment are obtained by
numerical integration of the conditional density kernel: the $u$-form
@eq-post-prod for production,

$$
h(u; y_i, a_i)
\;\propto\; \exp\!\bigl(-e^{a_i - u} - (y_i + b)\,u\bigr)\, u^{\alpha-1},
\qquad u > 0,
$$ {#eq-post-kernel}

with integration grid centred at the production-kernel mode
$\tilde{u}_i = a_i - \log(y_i + b)$, and the $v = e^{a_i + u}$ transform of
@eq-post-cost for cost (the same single-exponential damping used in the
cost-PMF integral @eq-cost-quad), with grid centred at
$v_i^* = \max(e^{a_i},\, y_i - b)$ and width set by an approximating Gamma
standard deviation $\sqrt{\max(y_i - b,\,1) + \alpha}$. An approximate 95%
conditional interval is constructed as
$\widehat{\mathrm{TE}}_i \pm 1.96\,\widehat{\sigma}_i$, where
$\widehat{\sigma}_i^2 = \mathrm{E}[e^{-2u_i} \mid y_i,\, \mathbf{x}_i] - \widehat{\mathrm{TE}}_i^2$
is estimated from the same integration. In the empirical application (patent
data, $n = 1730$), the closed-form series suffers catastrophic cancellation for
observations with $y_i > 100$ and $\exp(\hat{a}_i) > 200$; the quadrature-based
PMF evaluator described above is used as a fallback in these cases, with
accuracy verified to exceed $10^{-4}$ relative tolerance against
high-precision adaptive quadrature.

With the estimation framework established, @sec-simulations examines
finite-sample performance of the MLE across twelve DGP configurations that
vary the true $\alpha_0$, $b_0$, and sample size $n$.
