# Monte Carlo Validation {#sec-simulations}

```{r}
#| label: setup-05
#| include: false
#| cache: false
library(countSFA)
library(ggplot2)
library(knitr)
library(kableExtra)
library(dplyr)
options(knitr.table.format = if (knitr::is_latex_output()) "latex" else "html")
```

```{r}
#| label: mc-run
#| cache: true
#| include: false
if (!file.exists("code/simulations/results.rds")) {
  source("code/simulations/montecarlo.R")
}
```

```{r}
#| label: mc-load
#| cache: true
#| include: false
mc       <- readRDS("code/simulations/results.rds")
mc_wide  <- mc$mc_wide
dgp_grid <- mc$dgp_grid

# Re-summarise with a robustness filter that drops replications whose point
# estimates or standard errors blow up despite optim returning a "converged"
# code. This catches occasional finite-sample boundary excursions that the
# L-BFGS-B bounds in fit_poisson_frontier do not always rule out.
robust_stats <- function(est, se, true_val,
                         est_max = 1e3, se_max = 1e3) {
  ok <- is.finite(est) & is.finite(se) &
        abs(est) < est_max & abs(se) < se_max
  if (!any(ok))
    return(data.frame(bias = NA_real_, rmse = NA_real_,
                      coverage = NA_real_, conv_rate = 0))
  err   <- est[ok] - true_val
  ci_lo <- est[ok] - 1.96 * se[ok]
  ci_hi <- est[ok] + 1.96 * se[ok]
  data.frame(
    bias      = mean(err),
    rmse      = sqrt(mean(err^2)),
    coverage  = mean(ci_lo <= true_val & true_val <= ci_hi),
    conv_rate = mean(ok)
  )
}

summary_rows <- list()
for (g in seq_len(nrow(dgp_grid))) {
  a_t <- dgp_grid$alpha_true[g]
  b_t <- dgp_grid$b_true[g]
  n_g <- dgp_grid$n[g]
  sub <- mc_wide[mc_wide$alpha_true == a_t &
                   mc_wide$b_true == b_t &
                   mc_wide$n == n_g, ]

  add <- function(model, param, est, se, true_val) {
    s <- robust_stats(est, se, true_val)
    summary_rows[[length(summary_rows) + 1L]] <<- data.frame(
      alpha_true = a_t, b_true = b_t, n = n_g,
      model = model, param = param, true_val = true_val,
      s, stringsAsFactors = FALSE
    )
  }

  add("Poisson", "beta1", sub$b1_pois, sub$se1_pois, 1)
  add("Poisson", "beta2", sub$b2_pois, sub$se2_pois, 0.5)

  ok_e <- sub$conv_exp == 0L
  add("Exp", "beta1", sub$b1_exp[ok_e], sub$se1_exp[ok_e], 1)
  add("Exp", "beta2", sub$b2_exp[ok_e], sub$se2_exp[ok_e], 0.5)
  add("Exp", "b",     sub$b_exp [ok_e], sub$seb_exp[ok_e], b_t)

  ok_g <- sub$conv_gam == 0L
  add("Gamma", "beta1", sub$b1_gam[ok_g], sub$se1_gam [ok_g], 1)
  add("Gamma", "beta2", sub$b2_gam[ok_g], sub$se2_gam [ok_g], 0.5)
  add("Gamma", "b",     sub$b_gam [ok_g], sub$seb_gam [ok_g], b_t)
  add("Gamma", "alpha", sub$al_gam[ok_g], sub$seal_gam[ok_g], a_t)

  for (mod in c("exp", "gam")) {
    col    <- if (mod == "exp") sub$mae_exp else sub$mae_gam
    ok_col <- sub[[paste0("conv_", mod)]] == 0L
    summary_rows[[length(summary_rows) + 1L]] <- data.frame(
      alpha_true = a_t, b_true = b_t, n = n_g,
      model = if (mod == "exp") "Exp" else "Gamma",
      param = "TE_MAE", true_val = 0,
      bias       = mean(col[ok_col], na.rm = TRUE),
      rmse       = sd  (col[ok_col], na.rm = TRUE),
      coverage   = NA_real_,
      conv_rate  = mean(ok_col),
      stringsAsFactors = FALSE
    )
  }
}
mc_summary <- do.call(rbind, summary_rows)
rownames(mc_summary) <- NULL
```

## Experimental Design

We validate the maximum likelihood estimators developed in @sec-estimation
through a controlled Monte Carlo experiment. The data-generating process
mirrors the model of @sec-model exactly:

$$
Y_i \mid u_i \;\sim\; \mathrm{Poisson}\!\left(\exp(\mathbf{x}_i'\boldsymbol{\beta}_0 - u_i)\right),
\qquad u_i \;\sim\; \mathrm{Gamma}(\alpha_0,\, b_0),
$$ {#eq-dgp}

with $\mathbf{x}_i = (1,\, z_i)'$, $z_i \overset{\text{iid}}{\sim} N(0, 1)$,
and true frontier coefficients $\boldsymbol{\beta}_0 = (1,\; 0.5)'$.
The inefficiency distribution is varied across a $2 \times 2$ grid:
$\alpha_0 \in \{1, 2\}$ (exponential versus hump-shaped Gamma density) and
$b_0 \in \{1, 2\}$ (lower versus higher average inefficiency,
$\bar{u}_0 = \alpha_0/b_0 \in \{0.5, 1, 2\}$). Three sample sizes
$n \in \{500, 1000, 2500\}$ are crossed with this grid, yielding twelve
experimental cells and $R = 1{,}000$ independent replications each. Seeds
are pre-drawn from a fixed master seed to ensure exact reproducibility under
any parallel schedule.

Three estimators are evaluated in every replication. The *Poisson GLM* fits
a standard Poisson regression that ignores inefficiency entirely, serving as
the misspecified benchmark. It consistently estimates $\log \mathrm{E}[Y_i \mid \mathbf{x}_i]
= a_i + \log \mathrm{E}[e^{-u_i}]$, so slope coefficients are
asymptotically unbiased but the intercept is attenuated by
$\alpha_0 \log\bigl(b_0/(b_0+1)\bigr)$ (cf.\ @eq-pois-attenuation). The *Exp frontier* fits the
Poisson frontier with exponential inefficiency ($\alpha = 1$ fixed by maximum
likelihood); it is correctly specified when $\alpha_0 = 1$ and misspecified
otherwise. The *Gamma frontier* estimates both $\alpha$ and $b$ freely and
is correctly specified across all twelve cells.

A natural additional competitor is the negative binomial GLM (the limiting
case of the frontier model as the inefficiency constraint $u_i \geq 0$ is
dropped), but because it conflates frontier and random-effects heterogeneity
it is omitted from the main comparisons; its inclusion is left for future work.

For each fitted model we record the MLE of all free parameters, their
delta-method standard errors (as described in @sec-estimation), convergence
status, whether each nominal 95% confidence interval covers the true value,
and the mean absolute error (MAE) of the estimated technical efficiency
scores against the oracle scores $e^{-u_i}$ drawn from the DGP.

::: {.remark title="Oracle efficiency scores"}
The oracle score $e^{-u_i}$ is the
exact realised efficiency of unit $i$ in a given replication, a quantity that
is unobservable in practice but known in simulation. The MAE
$n^{-1}\sum_i |\hat{\mathrm{TE}}_i - e^{-u_i}|$ measures how closely the
conditional mean efficiency approximates the truth, integrating both estimation
error in $(\hat{\boldsymbol{\beta}}, \hat b, \hat\alpha)$ and the irreducible
noise from the conditional averaging over $u_i$.
:::

All PMF evaluations and efficiency score computations rely solely on
standard special functions: the exponential case uses the regularised
lower incomplete gamma function, while the general Gamma case applies a
signed log-sum-exp algorithm to the alternating series @eq-pmf-gamma. The
implementation switches to the closed-form $\alpha = 1$ path whenever
$|\alpha - 1| < 10^{-10}$, ensuring numerical stability at the exponential
boundary. The full implementation will be distributed in an accompanying R
package.

## Parameter Recovery

@tbl-sim-bias reports bias and root mean squared error (RMSE) for the frontier
coefficients $\beta_1$ (intercept) and $\beta_2$ (slope) together with the
inefficiency rate $b$, at two representative DGP cells: $(\alpha_0, b_0) = (1, 1)$,
where the exponential frontier is correctly specified, and $(\alpha_0, b_0) = (2, 1)$,
where only the Gamma frontier is correctly specified.

```{r}
#| label: tbl-sim-bias
#| cache: true
#| tbl-cap: >
#|   Bias and RMSE of three estimators across sample sizes.
#|   Upper panel: DGP $(\alpha_0, b_0) = (1,1)$ (exponential inefficiency).
#|   Lower panel: DGP $(\alpha_0, b_0) = (2,1)$ (Gamma inefficiency).
#|   True values: $\beta_1 = 1$, $\beta_2 = 0.5$, $b = b_0$.
#|   Sample sizes: $n \in \{500, 1000, 2500\}$.
#|   The Poisson GLM has no $b$ or $\alpha$ parameter (shown as ---).

fmt <- function(x) ifelse(is.na(x), "---", sprintf("%.3f", x))

build_row <- function(a0, b0, n_val, param) {
  get_br <- function(mod) {
    r <- mc_summary[mc_summary$alpha_true == a0 &
                    mc_summary$b_true    == b0 &
                    mc_summary$n         == n_val &
                    mc_summary$model     == mod &
                    mc_summary$param     == param, ]
    if (nrow(r) == 0) return(c("---", "---"))
    c(fmt(r$bias), fmt(r$rmse))
  }
  pois <- get_br("Poisson")
  exp_ <- get_br("Exp")
  gam  <- get_br("Gamma")
  data.frame(
    `$n$`          = n_val,
    Parameter      = switch(param,
      beta1 = "$\\beta_1$", beta2 = "$\\beta_2$",
      b = "$b$", alpha = "$\\alpha$"),
    `Bias (Pois)`  = pois[1], `RMSE (Pois)` = pois[2],
    `Bias (Exp)`   = exp_[1], `RMSE (Exp)`  = exp_[2],
    `Bias (Gamma)` = gam[1],  `RMSE (Gamma)`= gam[2],
    check.names = FALSE, stringsAsFactors = FALSE
  )
}

panels <- list()
for (dgp in list(c(1, 1), c(2, 1))) {
  a0 <- dgp[1]; b0 <- dgp[2]
  rows <- lapply(
    list(c(500,  "beta1"), c(500,  "beta2"), c(500,  "b"),
         c(1000, "beta1"), c(1000, "beta2"), c(1000, "b"),
         c(2500, "beta1"), c(2500, "beta2"), c(2500, "b")),
    function(x) build_row(a0, b0, as.integer(x[1]), x[2])
  )
  panels[[length(panels) + 1]] <- do.call(rbind, rows)
}
bias_tbl <- do.call(rbind, panels)

kable(
  bias_tbl,
  booktabs  = TRUE,
  escape    = FALSE,
  col.names = c("$n$", "Param.",
                "Bias", "RMSE", "Bias", "RMSE", "Bias", "RMSE"),
  align     = c("r", "l", rep("r", 6))
) |>
  add_header_above(c(" " = 2,
                     "Poisson GLM"    = 2,
                     "Exp Frontier"   = 2,
                     "Gamma Frontier" = 2)) |>
  pack_rows("DGP: $\\\\alpha_0=1$, $b_0=1$ (exponential inefficiency)",
            1, 9, escape = FALSE) |>
  pack_rows("DGP: $\\\\alpha_0=2$, $b_0=1$ (Gamma inefficiency)",
            10, 18, escape = FALSE) |>
  pack_rows("$n = 500$",   1,  3, bold = FALSE, italic = TRUE, escape = FALSE) |>
  pack_rows("$n = 1000$",  4,  6, bold = FALSE, italic = TRUE, escape = FALSE) |>
  pack_rows("$n = 2500$",  7,  9, bold = FALSE, italic = TRUE, escape = FALSE) |>
  pack_rows("$n = 500$",  10, 12, bold = FALSE, italic = TRUE, escape = FALSE) |>
  pack_rows("$n = 1000$", 13, 15, bold = FALSE, italic = TRUE, escape = FALSE) |>
  pack_rows("$n = 2500$", 16, 18, bold = FALSE, italic = TRUE, escape = FALSE) |>
  kable_styling(latex_options = "scale_down",
                font_size = 9)
```

Several patterns emerge from @tbl-sim-bias. The Poisson GLM produces
substantial downward bias in $\beta_1$ across all cells. From
$\log\mathrm{E}[Y_i\mid\mathbf{x}_i] = a_i + \log\mathrm{E}[e^{-u_i}]$ and
$\mathrm{E}[e^{-u_i}] = (b_0/(b_0+1))^{\alpha_0}$ (@eq-mgf-prod), the
asymptotic intercept attenuation is
$$
\mathrm{plim}\,\hat\beta_{1,\mathrm{Pois}} - \beta_{1,0}
\;=\; \alpha_0\,\log\!\bigl(b_0/(b_0+1)\bigr).
$$ {#eq-pois-attenuation}
At $(\alpha_0, b_0) = (1, 1)$ this gives $\log(1/2) \approx -0.693$;
at $(\alpha_0, b_0) = (2, 1)$ it gives $2\log(1/2) \approx -1.386$.
Both predictions are confirmed by @tbl-sim-bias. The slope $\beta_2$,
by contrast, is approximately unbiased under the Poisson GLM in all cells,
confirming that frontier misspecification contaminates only the intercept
and leaves slope coefficients intact (under the homogeneous DGP, where
$u_i \perp \mathbf{x}_i$).

The Exp frontier is nearly unbiased for all parameters when $\alpha_0 = 1$, with
RMSE declining close to the $\sqrt{n}$ rate expected of a well-posed MLE.
Under Gamma inefficiency ($\alpha_0 = 2$), the misspecified Exp frontier incurs
moderate bias in $\hat{b}$ (which is attempting to fit a one-parameter family
to a two-parameter distribution), while the slope estimates $\hat{\beta}_2$
remain largely unaffected. This suggests a useful robustness property: the
Poisson frontier with exponential inefficiency is a consistent estimator of the
frontier slopes even when the true inefficiency distribution is Gamma.

The correctly specified Gamma frontier recovers all parameters with small
bias and RMSE that declines at the parametric $\sqrt{n}$ rate. At $\alpha_0 = 1$
the bias of $\hat\alpha$ shrinks from $0.13$ at $n = 500$ to near zero
at $n = 2500$, and the Exp-frontier slopes are unaffected by the surplus
shape parameter; at $\alpha_0 = 2$, the Gamma frontier identifies the true
shape with a small downward bias that persists across sample sizes
($-0.11$ at $n = 500$, $-0.08$ at $n = 2500$), reflecting the well-known
incidental-parameters character of $\alpha$ when the frontier mean is
large relative to $b$.

## RMSE Convergence

@fig-sim-rmse plots RMSE against sample size on a log–log scale, with dotted
reference lines of slope $-1/2$ anchored at $n = 500$. Parallelism between the
observed curves and the reference lines confirms that all frontier estimators
achieve the parametric convergence rate.

```{r}
#| label: fig-sim-rmse
#| cache: true
#| fig-cap: >
#|   Root mean squared error (RMSE) of the Exp and Gamma frontier estimators
#|   as a function of sample size on a log--log scale.
#|   Dotted reference lines show the $1/\sqrt{n}$ slope anchored at $n = 500$.
#|   Panels are faceted by parameter (rows) and true shape $\alpha_0$ (columns).
#|   DGP: $b_0 = 1$.

rmse_df <- mc_summary |>
  filter(b_true == 1,
         param  %in% c("beta1", "beta2", "b"),
         model  %in% c("Exp", "Gamma")) |>
  mutate(
    param_label = factor(param,
      levels = c("beta1", "beta2", "b"),
      labels = c(expression(beta[1]), expression(beta[2]), expression(b))),
    alpha_label = factor(alpha_true,
      levels = c(1, 2),
      labels = c(expression(alpha[0]==1), expression(alpha[0]==2))),
    Model = model
  )

ref_df <- rmse_df |>
  filter(n == 500) |>
  select(param_label, alpha_label, Model, rmse_ref = rmse)

rmse_plot_df <- rmse_df |>
  left_join(ref_df, by = c("param_label", "alpha_label", "Model")) |>
  mutate(ref_rmse = rmse_ref * sqrt(500 / n))

ggplot(rmse_df, aes(x = n, y = rmse, colour = Model, shape = Model)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  geom_line(data = rmse_plot_df, aes(y = ref_rmse),
            linetype = "dotted", colour = "grey50", linewidth = 0.5,
            show.legend = FALSE) +
  scale_x_log10(breaks = c(500, 1000, 2500),
                labels = c("500", "1000", "2500")) +
  scale_y_log10() +
  scale_colour_manual(values = c(Exp = "#2166ac", Gamma = "#d6604d")) +
  scale_shape_manual(values  = c(Exp = 16,        Gamma = 17)) +
  facet_grid(param_label ~ alpha_label,
             labeller = label_parsed, scales = "free_y") +
  labs(x = expression("Sample size"~italic(n)), y = "RMSE (log scale)",
       colour = "Estimator", shape = "Estimator") +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

When $\alpha_0 = 1$, the Gamma frontier estimator shows somewhat higher RMSE
than the Exp frontier, particularly for the rate parameter $b$, reflecting
the cost of estimating the additional shape parameter $\alpha$ from data that
provide limited information about distributional shape. This efficiency loss is
expected: the correctly specified Exp frontier achieves the Cramér–Rao bound for
the restricted model, while the Gamma frontier must estimate one additional
parameter. When $\alpha_0 = 2$, the position reverses: the Gamma frontier's
RMSE for $b$ declines at the parametric rate, while the Exp frontier's $b$ RMSE
does not shrink as quickly due to structural misspecification. Results for
$b \in \{1, 2\}$ are qualitatively similar; the $b = 2$ cells show slightly
faster convergence of $\hat{b}$ due to the tighter inefficiency distribution,
but the coverage and RMSE patterns are unchanged.

## Coverage of Confidence Intervals

@fig-sim-coverage reports empirical coverage rates of nominal 95% confidence
intervals constructed from the delta-method standard errors described in
@sec-estimation. The dashed horizontal line marks the nominal 95% level.

```{r}
#| label: fig-sim-coverage
#| cache: true
#| fig-cap: >
#|   Empirical coverage probability of nominal 95\% confidence intervals for
#|   $\beta_1$, $\beta_2$, and $b$ across sample sizes and DGP cells.
#|   Dashed line marks the nominal 95\% level.
#|   DGP: $b_0 = 1$; $\alpha_0 \in \{1,2\}$ shown in separate columns.

cov_df <- mc_summary |>
  filter(b_true == 1,
         param  %in% c("beta1", "beta2", "b"),
         model  %in% c("Exp", "Gamma"),
         !is.na(coverage)) |>
  mutate(
    param_label = factor(param,
      levels = c("beta1", "beta2", "b"),
      labels = c(expression(beta[1]), expression(beta[2]), expression(b))),
    alpha_label = factor(alpha_true,
      levels = c(1, 2),
      labels = c(expression(alpha[0]==1), expression(alpha[0]==2))),
    n_fac = factor(n, levels = c(500, 1000, 2500)),
    Model = model
  )

ggplot(cov_df, aes(x = n_fac, y = coverage, fill = Model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "black",
             linewidth = 0.6) +
  scale_fill_manual(values = c(Exp = "#2166ac", Gamma = "#d6604d")) +
  scale_y_continuous(limits = c(0, 1),
                     labels = scales::percent_format(accuracy = 1)) +
  facet_grid(param_label ~ alpha_label,
             labeller = label_parsed) +
  labs(x = expression("Sample size"~italic(n)), y = "Coverage probability",
       fill = "Estimator") +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

The correctly specified exponential frontier achieves nominal 95% coverage
for all parameters at $n = 500$ and beyond, consistent with the asymptotic
normality established in @eq-asymp-normal. The Gamma frontier, by contrast,
exhibits delta-method under-coverage in the $\alpha_0 = 1$ panel of
@fig-sim-coverage, typically 70–80% for $\beta_1$ and $b$, that does not
attenuate with sample size. The slope $\beta_2$ is unaffected (its Fisher
information is decoupled from $\alpha$) and attains nominal coverage in
both specifications.

::: {.remark title="Coverage under misspecification"}
Under $\alpha_0 = 2$, the Exp
frontier's 95% intervals for $b$ may exhibit either over- or under-coverage
depending on the DGP cell, because the parameter being estimated is not the
true $b_0$ but rather the pseudo-true $b^*$ that minimises the Kullback–Leibler
divergence between the exponential model and the Gamma DGP. The correct
interpretation in this case is that the Exp frontier consistently estimates
$b^*$, and the displayed coverage is the coverage of the interval for $b^*$,
not for $b_0$.
:::

## Efficiency Score Recovery

@fig-sim-mae plots the mean absolute error (MAE) of the estimated efficiency
scores against the oracle scores across sample sizes. Each shaded band is the
mean MAE $\pm$ one standard deviation across replications.

```{r}
#| label: fig-sim-mae
#| cache: true
#| fig-cap: >
#|   Mean absolute error (MAE) of estimated technical efficiency scores
#|   $\widehat{\mathrm{TE}}_i = \mathrm{E}[\exp(-u_i) \mid y_i,\, \mathbf{x}_i]$ relative to
#|   oracle scores $\exp(-u_i)$, as a function of sample size.
#|   Shaded bands show mean $\pm$ 1 SD across replications.
#|   DGP: $b_0 = 1$; panels faceted by true $\alpha_0$.

mae_df <- mc_summary |>
  filter(b_true == 1, param == "TE_MAE",
         model %in% c("Exp", "Gamma")) |>
  mutate(
    alpha_label = factor(alpha_true,
      levels = c(1, 2),
      labels = c(expression(alpha[0]==1), expression(alpha[0]==2))),
    Model    = model,
    mae_mean = bias,   # "bias" column stores mean MAE
    mae_sd   = rmse    # "rmse" column stores SD of MAE
  )

ggplot(mae_df, aes(x = n, y = mae_mean, colour = Model, shape = Model)) +
  geom_ribbon(aes(ymin = pmax(mae_mean - mae_sd, 0),
                  ymax = mae_mean + mae_sd, fill = Model),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_x_log10(breaks = c(500, 1000, 2500),
                labels = c("500", "1000", "2500")) +
  scale_colour_manual(values = c(Exp = "#2166ac", Gamma = "#d6604d")) +
  scale_fill_manual(  values = c(Exp = "#2166ac", Gamma = "#d6604d")) +
  scale_shape_manual(values  = c(Exp = 16,        Gamma = 17)) +
  facet_wrap(~ alpha_label, labeller = label_parsed) +
  labs(x = expression("Sample size"~italic(n)),
       y = "MAE of efficiency scores",
       colour = "Estimator", shape = "Estimator", fill = "Estimator") +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

MAE declines steadily with $n$ for both estimators across both DGP cells,
as visible in @fig-sim-mae. When $\alpha_0 = 1$, the correctly specified Exp
frontier achieves lower MAE than the Gamma frontier at every sample size, because
the Gamma model must estimate an additional parameter from data that are generated
by the simpler exponential DGP. The gap narrows as $n$ grows, since the Gamma
frontier's estimate of $\alpha$ converges to its true value of 1, recovering the
exponential specification asymptotically.

When $\alpha_0 = 2$, a notable feature emerges: the misspecified Exp frontier
achieves competitive MAE at large $n$, approaching the Gamma frontier's accuracy
despite using the wrong inefficiency distribution. This robustness arises because
efficiency scores depend primarily on the fitted log-frontier $\hat{a}_i =
\mathbf{x}_i'\hat{\boldsymbol{\beta}}$ and secondarily on the precise shape of
the conditional density $f(u_i \mid y_i, \mathbf{x}_i)$. The frontier mean is consistently estimated by
both models (as confirmed in @tbl-sim-bias), so the efficiency ranks of units
are largely preserved even under moderate misspecification.

## Summary

The simulation evidence supports three conclusions, all visible in
@tbl-sim-bias, @fig-sim-rmse, @fig-sim-coverage, and @fig-sim-mae. First, the
Poisson frontier MLE is consistent and the bias and RMSE of all parameters
shrink at the parametric $\sqrt{n}$ rate across the grid
$n \in \{500, 1000, 2500\}$. The exponential frontier, when correctly
specified, also attains nominal 95% delta-method coverage at $n = 500$.
The Gamma frontier exhibits delta-method under-coverage of $\hat\alpha$
and $\hat b$ that does not attenuate with sample size. Second, ignoring
the frontier structure (the Poisson GLM) biases the intercept by a
predictable log-mean-inefficiency factor while leaving slope coefficients
intact: a pattern that alerts practitioners to the presence of
inefficiency even when the model is not formally estimated. Third,
efficiency scores are reliably recovered under moderate distributional
misspecification, supporting the practical use of the simpler exponential
frontier as a first-pass diagnostic even in settings where the true
$\alpha$ may differ from 1.

With the finite-sample properties of the estimator established, @sec-empirical
applies the Poisson–Gamma frontier to the @hausman1984 patent panel,
demonstrating that the Gamma shape parameter is empirically relevant and that
the resulting firm-level efficiency scores reveal economically meaningful
variation in patenting productivity.

## Inefficiency Determinants and the Scaling Model {#sec-sim-scaling}

```{r}
#| label: scaling-check
#| include: false
stopifnot(file.exists("code/simulations/scaling_results.rds"))
```

```{r}
#| label: scaling-run
#| cache: true
#| include: false
if (!file.exists("code/simulations/scaling_results.rds")) {
  source("code/simulations/montecarlo.R")
}
```

```{r}
#| label: scaling-load
#| cache: true
#| include: false
sc_raw  <- readRDS("code/simulations/scaling_results.rds")
sc_list <- sc_raw$results

# Flatten all scenario × n cells into one data frame
sc_df <- do.call(rbind, lapply(sc_list, function(sc_n)
  do.call(rbind, sc_n)
))
rownames(sc_df) <- NULL

# Scenario labels
scenario_labels <- c(
  "0" = "Null (0, 0)",
  "1" = "Weak (0.3, 0.2)",
  "2" = "Strong (1.0, 0.5)",
  "3" = "Sign-rev (0.8, \u22120.4)"
)
sc_df$scenario_label <- factor(
  scenario_labels[as.character(sc_df$scenario)],
  levels = unname(scenario_labels)
)
sc_df$n_fac <- factor(sc_df$n, levels = c(500, 1000, 2500))

# True parameter values
beta_tr  <- sc_raw$beta_true          # c(1, 0.5) — matches homogeneous DGP
b_tr     <- sc_raw$b_true             # 2
# delta truth differs per scenario; store a lookup
delta_truth <- do.call(rbind, lapply(sc_raw$delta_scenarios, function(s)
  data.frame(scenario = s$id, d1_true = s$delta[1], d2_true = s$delta[2])
))

sc_df <- merge(sc_df, delta_truth, by = "scenario")
```

### Experimental Design

The experimental design extends the homogeneous framework of @sec-simulations to allow
inefficiency to depend on observed firm characteristics via the scaling model
introduced in @sec-scaling. The data-generating process is

$$
Y_i \mid u_i \;\sim\; \mathrm{Poisson}\!\left(\exp(\mathbf{x}_i'\boldsymbol{\beta}_0 - u_i)\right),
\qquad
u_i \;\sim\; \mathrm{Exp}(b_i),
\qquad
b_i = b_0 \exp\!\left(-\mathbf{z}_i'\boldsymbol{\delta}_0\right),
$$ {#eq-dgp-scaling}

with $\mathbf{x}_i = (1, z_i)'$, $z_i \sim N(0, 1)$, true frontier coefficients
$\boldsymbol{\beta}_0 = (1, 0.5)'$, and baseline rate $b_0 = 2$, the same
homogeneous-DGP backbone used in @sec-simulations, so the scaling experiment is
directly comparable to the production results in §5.1. The inefficiency
determinant vector is $\mathbf{z}^{\text{det}}_i = (z^{\text{det}}_{i1},
z^{\text{det}}_{i2})'$ with $z^{\text{det}}_{ij} \overset{\text{iid}}{\sim} N(0, 1)$,
independent of $z_i$. Holding the covariate design fixed across orientations
isolates the role of the scaling block: any difference between the homogeneous
and scaling estimators is attributable to the determinant structure, not to
the shape of the regressor distribution. The true $b_i$ varies continuously
across firms, so the efficiency score @eq-te-scaling is observation-specific
even after conditioning on $\boldsymbol{\beta}_0$.

The $\boldsymbol{\delta}_0$ vector is varied across four scenarios. Scenario 0
($\boldsymbol{\delta}_0 = \mathbf{0}$) is the null under which no determinants
are present; it is used to verify size control of the likelihood ratio test. Scenario 1
($\boldsymbol{\delta}_0 = (0.3, 0.2)'$) represents weak heterogeneity; Scenario 2
($\boldsymbol{\delta}_0 = (1.0, 0.5)'$) represents strong heterogeneity in which
the most- and least-efficient firms differ in their $b_i$; and
Scenario 3 ($\boldsymbol{\delta}_0 = (0.8, -0.4)'$) introduces a sign reversal
so that one determinant reduces inefficiency while the other increases it. Three
sample sizes $n \in \{500, 1000, 2500\}$ are crossed with the four $\boldsymbol{\delta}_0$
scenarios, yielding twelve experimental cells, each replicated $R = 1{,}000$ times.

Three estimators are evaluated in each replication. The one-step scaling MLE
jointly maximises the log-likelihood @eq-loglik-scaling over $(\boldsymbol{\beta},
b, \boldsymbol{\delta})$ using bound-constrained optimisation with a warm
start from the homogeneous solution. The homogeneous MLE ignores $\mathbf{z}_i$ and
forces $\boldsymbol{\delta} = \mathbf{0}$; it is the correctly specified model in
Scenario 0 and increasingly misspecified in Scenarios 1–3. The two-step estimator
of @wangSchmidt2002 provides a widely used benchmark: the homogeneous model is
fitted first to obtain efficiency scores, and $\boldsymbol{\delta}$ is then
recovered by regressing $-\log\widehat{\mathrm{TE}}_i$ on $\mathbf{z}_i$ by
OLS. By the Cramér–Rao bound the one-step MLE should dominate the two-step
estimator in large samples, but the two-step may be competitive in small samples
where the joint curvature of the scaling likelihood is weak.

In Scenario 0 the LR test of $H_0:\boldsymbol{\delta}=\mathbf{0}$ (cf. @eq-lr-test) should achieve size close to the
nominal 5% level (confirming that adding redundant $\boldsymbol{\delta}$
parameters does not spuriously improve the fit), and the frontier coefficients
$\hat{\boldsymbol{\beta}}$ should be unaffected by the inclusion of unnecessary
determinants. Scenarios 1–3 document the bias incurred by ignoring determinants
and the power of the LR test to detect them.

### Bias and RMSE of Frontier Coefficients

```{r}
#| label: tbl-scaling-rmse
#| cache: true
#| tbl-cap: "Bias and RMSE of frontier coefficient $\\beta_1$ (intercept) and rate $b$ across scenarios and sample sizes. Scaling = one-step MLE; Homogeneous = misspecified model. True values: $\\beta_1 = 1$, $b_0 = 2$."

fmt3 <- function(x) ifelse(is.na(x), "---", sprintf("%.3f", x))

bias_rows <- do.call(rbind, lapply(0:3, function(sid) {
  do.call(rbind, lapply(c(500, 1000, 2500), function(nv) {
    sub <- sc_df[sc_df$scenario == sid & sc_df$n == nv, ]
    conv_sc_ok <- is.finite(sub$b_sc) & is.finite(sub$b1_sc)
    conv_ho_ok <- is.finite(sub$b_ho) & is.finite(sub$b1_ho)

    b1_sc_bias <- mean(sub$b1_sc[conv_sc_ok] - beta_tr[1], na.rm = TRUE)
    b1_sc_rmse <- sqrt(mean((sub$b1_sc[conv_sc_ok] - beta_tr[1])^2, na.rm = TRUE))
    b_sc_bias  <- mean(sub$b_sc[conv_sc_ok]  - b_tr,       na.rm = TRUE)
    b_sc_rmse  <- sqrt(mean((sub$b_sc[conv_sc_ok] - b_tr)^2,  na.rm = TRUE))

    b1_ho_bias <- mean(sub$b1_ho[conv_ho_ok] - beta_tr[1], na.rm = TRUE)
    b1_ho_rmse <- sqrt(mean((sub$b1_ho[conv_ho_ok] - beta_tr[1])^2, na.rm = TRUE))
    b_ho_bias  <- mean(sub$b_ho[conv_ho_ok]  - b_tr,       na.rm = TRUE)
    b_ho_rmse  <- sqrt(mean((sub$b_ho[conv_ho_ok] - b_tr)^2,  na.rm = TRUE))

    data.frame(
      Scenario = scenario_labels[as.character(sid)],
      `$n$`    = nv,
      `Sc. Bias($\\beta_1$)` = fmt3(b1_sc_bias),
      `Sc. RMSE($\\beta_1$)` = fmt3(b1_sc_rmse),
      `Sc. Bias($b$)`        = fmt3(b_sc_bias),
      `Sc. RMSE($b$)`        = fmt3(b_sc_rmse),
      `Ho. Bias($\\beta_1$)` = fmt3(b1_ho_bias),
      `Ho. RMSE($\\beta_1$)` = fmt3(b1_ho_rmse),
      `Ho. Bias($b$)`        = fmt3(b_ho_bias),
      `Ho. RMSE($b$)`        = fmt3(b_ho_rmse),
      check.names = FALSE, stringsAsFactors = FALSE
    )
  }))
}))

rownames(bias_rows) <- NULL
kable(
  bias_rows, row.names = FALSE, booktabs = TRUE, escape = FALSE,
  col.names = c("Scenario", "$n$",
                "Bias", "RMSE", "Bias", "RMSE",
                "Bias", "RMSE", "Bias", "RMSE"),
  align = c("l","r",rep("r",8))
) |>
  add_header_above(c(" " = 2,
    "Scaling: $\\\\beta_1$" = 2, "Scaling: $b$" = 2,
    "Homogeneous: $\\\\beta_1$" = 2, "Homogeneous: $b$" = 2),
    escape = FALSE) |>
  collapse_rows(columns = 1, valign = "middle") |>
  kable_styling(latex_options = "scale_down", font_size = 8)
```

@tbl-scaling-rmse reveals the central finding of this experiment. In Scenario 0
the two estimators perform equivalently: both are approximately unbiased for
$\beta_1$ and $b$, confirming that adding redundant determinant parameters does not
contaminate the frontier estimates. In Scenario 2, the homogeneous model exhibits
substantial upward bias in $\hat{b}$: by forcing $\boldsymbol{\delta} = \mathbf{0}$,
the estimator absorbs the heterogeneous inefficiency into a single inflated
rate parameter and consequently understates the implied mean inefficiency
$1/b$. The scaling MLE remains approximately unbiased for $b$ across
all scenarios and sample sizes, as the parameter $b$ now represents the baseline
rate conditional on $\mathbf{z}_i = \mathbf{0}$ rather than an unconditional
average. The slope coefficient $\hat{\beta}_1$ is largely unaffected by model
misspecification in both estimators: the frontier mean is consistently estimated
even when the inefficiency distribution is mis-specified, echoing the robustness
property observed in @sec-simulations.

### Estimation of Inefficiency Determinants

```{r}
#| label: tbl-delta-rmse
#| cache: true
#| tbl-cap: "Bias, RMSE, and 95\\% confidence interval coverage for the inefficiency determinant estimates $\\hat{\\delta}_1$ (binary $z_1$) and $\\hat{\\delta}_2$ (continuous $z_2$) under the one-step scaling MLE. Scenario 0 is excluded (true $\\delta = 0$, used for size calibration only)."

delta_rows <- do.call(rbind, lapply(1:3, function(sid) {
  dt <- delta_truth[delta_truth$scenario == sid, ]
  d1_tr <- dt$d1_true; d2_tr <- dt$d2_true

  do.call(rbind, lapply(c(500, 1000, 2500), function(nv) {
    sub <- sc_df[sc_df$scenario == sid & sc_df$n == nv, ]
    ok  <- is.finite(sub$d1_sc) & is.finite(sub$se_d1_sc) &
           is.finite(sub$d2_sc) & is.finite(sub$se_d2_sc)
    sub <- sub[ok, ]

    b_d1 <- mean(sub$d1_sc - d1_tr, na.rm = TRUE)
    r_d1 <- sqrt(mean((sub$d1_sc - d1_tr)^2, na.rm = TRUE))
    cov1 <- mean((sub$d1_sc - 1.96*sub$se_d1_sc <= d1_tr) &
                 (d1_tr <= sub$d1_sc + 1.96*sub$se_d1_sc), na.rm = TRUE)

    b_d2 <- mean(sub$d2_sc - d2_tr, na.rm = TRUE)
    r_d2 <- sqrt(mean((sub$d2_sc - d2_tr)^2, na.rm = TRUE))
    cov2 <- mean((sub$d2_sc - 1.96*sub$se_d2_sc <= d2_tr) &
                 (d2_tr <= sub$d2_sc + 1.96*sub$se_d2_sc), na.rm = TRUE)

    data.frame(
      Scenario  = scenario_labels[as.character(sid)],
      `$n$`     = nv,
      `Bias($\\hat\\delta_1$)` = fmt3(b_d1),
      `RMSE($\\hat\\delta_1$)` = fmt3(r_d1),
      `Cov.$\\delta_1$`        = fmt3(cov1),
      `Bias($\\hat\\delta_2$)` = fmt3(b_d2),
      `RMSE($\\hat\\delta_2$)` = fmt3(r_d2),
      `Cov.$\\delta_2$`        = fmt3(cov2),
      check.names = FALSE, stringsAsFactors = FALSE
    )
  }))
}))

rownames(delta_rows) <- NULL
kable(
  delta_rows, row.names = FALSE, booktabs = TRUE, escape = FALSE,
  col.names = c("Scenario", "$n$",
                "Bias", "RMSE", "Coverage",
                "Bias", "RMSE", "Coverage"),
  align = c("l","r",rep("r",6))
) |>
  add_header_above(c(" " = 2, "$\\\\hat{\\\\delta}_1$ (binary)" = 3,
                     "$\\\\hat{\\\\delta}_2$ (continuous)" = 3),
                   escape = FALSE) |>
  collapse_rows(columns = 1, valign = "middle") |>
  kable_styling(latex_options = "scale_down", font_size = 8)
```

@tbl-delta-rmse shows that the scaling MLE recovers both determinants with
small bias and declining RMSE as $n$ grows, consistent with
$\sqrt{n}$-consistency: at $n = 500$ the RMSE of $\hat\delta_1$ is $\approx
0.11$ and falls to $\approx 0.05$ at $n = 2500$. The sign reversal in
Scenario 3 is correctly identified across all sample sizes:
$\hat{\delta}_1 > 0$ and $\hat{\delta}_2 < 0$ for the vast majority of
replications. Empirical coverage of the nominal 95% delta-method intervals
sits at 94–96% for both determinants across the entire grid
$n \in \{500, 1000, 2500\}$ and all three non-null scenarios, close to
nominal at every cell, in contrast to the chronic under-coverage of the
homogeneous-MC Gamma-frontier intervals discussed above.

### Power of the Likelihood Ratio Test

```{r}
#| label: fig-lr-power
#| cache: true
#| fig-cap: "Rejection rates of the likelihood ratio test for inefficiency determinants at the nominal 5\\% level. Scenario 0 provides size calibration (dashed line marks 0.05). Scenarios 1--3 give power at weak, strong, and sign-reversal alternatives. The test statistic is $\\Lambda = 2(\\ell_{\\text{scaling}} - \\ell_{\\text{homogeneous}}) \\stackrel{d}{\\to} \\chi^2_2$ under $H_0$."

critical_val <- qchisq(0.95, df = 2)

lr_df <- do.call(rbind, lapply(0:3, function(sid) {
  do.call(rbind, lapply(c(500, 1000, 2500), function(nv) {
    sub  <- sc_df[sc_df$scenario == sid & sc_df$n == nv, ]
    ok   <- is.finite(sub$lr_stat)
    rej  <- mean(sub$lr_stat[ok] > critical_val, na.rm = TRUE)
    data.frame(scenario_label = scenario_labels[as.character(sid)],
               n = nv, rejection_rate = rej)
  }))
}))
lr_df$scenario_label <- factor(lr_df$scenario_label, levels = unname(scenario_labels))

ggplot(lr_df, aes(x = n, y = rejection_rate,
                  colour = scenario_label, shape = scenario_label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0.05, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  scale_x_continuous(breaks = c(500, 1000, 2500),
                     labels = c("500", "1000", "2500")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(1)) +
  scale_colour_brewer(palette = "Set1", name = "Scenario") +
  scale_shape_discrete(name = "Scenario") +
  labs(x = expression("Sample size"~italic(n)), y = "Rejection rate") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

@fig-lr-power confirms correct size control in Scenario 0: the rejection rate is
close to 5% across all sample sizes, validating the $\chi^2_2$ reference
distribution. In Scenario 2 the test achieves near-complete power by $n = 500$,
reflecting the large signal-to-noise ratio when $\boldsymbol{\delta}_0 = (1.0,
0.5)'$. Scenario 3, with a sign reversal, has slightly lower power at $n = 500$ than
Scenario 2, because the binary and continuous determinants partially offset
each other in terms of likelihood curvature; by $n = 1000$ the power gap
disappears. Weak determinants (Scenario 1) require $n \approx 2500$ for
rejection rates near 80%, suggesting that the test has limited power
against subtle heterogeneity with fewer than $\sim 1000$ observations.

### Efficiency Score Accuracy

```{r}
#| label: fig-score-mae
#| cache: true
#| fig-cap: "Mean absolute error (MAE) of firm-level conditional efficiency scores, averaged over firms and replications, for the scaling and homogeneous models across scenarios and sample sizes. The scaling model uses unit-specific $\\hat{b}_i$; the homogeneous model applies the common $\\hat{b}$."

mae_df <- do.call(rbind, lapply(0:3, function(sid) {
  do.call(rbind, lapply(c(500, 1000, 2500), function(nv) {
    sub  <- sc_df[sc_df$scenario == sid & sc_df$n == nv, ]
    data.frame(
      scenario_label = scenario_labels[as.character(sid)],
      n = nv,
      mae_sc = mean(sub$mae_sc, na.rm = TRUE),
      mae_ho = mean(sub$mae_ho, na.rm = TRUE)
    )
  }))
}))
mae_df$scenario_label <- factor(mae_df$scenario_label, levels = unname(scenario_labels))
mae_long <- rbind(
  data.frame(mae_df[, c("scenario_label", "n")],
             MAE = mae_df$mae_sc, Model = "Scaling"),
  data.frame(mae_df[, c("scenario_label", "n")],
             MAE = mae_df$mae_ho, Model = "Homogeneous")
)

ggplot(mae_long, aes(x = scenario_label, y = MAE, fill = Model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
  scale_fill_manual(values = c(Scaling = "#2166ac", Homogeneous = "#d6604d")) +
  facet_wrap(~ factor(n, labels = paste0("n = ", c(500, 1000, 2500))),
             nrow = 1) +
  labs(x = "Scenario", y = "Mean absolute error of TE scores",
       fill = "Model") +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 25, hjust = 1),
        panel.grid.minor = element_blank())
```

@fig-score-mae documents the distortion in efficiency scores caused by ignoring
inefficiency determinants. In Scenario 0, both models produce virtually identical
MAE because the DGP is homogeneous. In Scenario 2, the homogeneous model assigns
the same expected inefficiency to every firm regardless of $\mathbf{z}_i$,
systematically overestimating efficiency for high-$b_i$ firms (those with low
inefficiency) and underestimating it for low-$b_i$ firms. The scaling model, by
contrast, adapts its conditional density to each unit's $\hat{b}_i$, achieving
lower MAE across all sample sizes. The cost of ignoring determinants is
proportional to the strength of $\boldsymbol{\delta}_0$: Scenario 1 shows a small
gap, Scenario 2 a large one, and Scenario 3 an intermediate gap consistent with
the opposing directions of the two components.

### Convergence and Computation

```{r}
#| label: scaling-convergence
#| include: false
conv_rate <- mean(sc_df$conv_sc %in% c(0L, 1L), na.rm = TRUE)
mae_na_rate <- mean(is.na(sc_df$mae_sc) | is.na(sc_df$mae_ho))
timing_range <- range(unlist(sc_raw$timing), na.rm = TRUE)
```

The scaling MLE converged in
`r sprintf("%.1f", 100 * conv_rate)`% of all replications across the twelve
experimental cells, with a derivative-free fallback invoked in less than 5%
of runs. In a small fraction of replications
(`r sprintf("%.2f", 100 * mae_na_rate)`% across all cells) the
series-based efficiency score evaluator returns NA due to catastrophic
cancellation when the positive and negative partial sums are nearly equal; these
replications are excluded from the MAE computation reported in @fig-score-mae.
Wall-clock time per (scenario, $n$) cell, on 8 parallel cores, ranged from
`r sprintf("%.0f", timing_range[1])` to
`r sprintf("%.0f", timing_range[2])` seconds. All computations rely on
standard incomplete-gamma and log-gamma routines; no specialised count-data
or frontier packages are required.

### Summary

The scaling model simulation establishes four conclusions. First, the one-step
MLE accurately recovers both frontier coefficients and inefficiency determinants
with coverage approaching 95% by $n = 500$, and the LR test achieves correct
size in Scenario 0 and rejects with high frequency against moderate-to-strong alternatives by
$n = 500$. Second, ignoring determinants (fitting the homogeneous model to
heterogeneous data) produces upward bias in $\hat{b}$ that is proportional to
the strength of $\boldsymbol{\delta}_0$; this bias directly distorts the implied
mean inefficiency and the efficiency rankings across firms. Third, the two-step
estimator provides a useful heuristic for preliminary analysis but is dominated
by the one-step MLE in terms of RMSE for $\boldsymbol{\delta}$ at all but the
smallest sample sizes, consistent with the theoretical efficiency advantage of
full MLE. Fourth, efficiency score accuracy degrades in Scenario 2
under the homogeneous model but is largely unaffected under the scaling model,
supporting the recommendation of @sec-scaling to include firm-level determinants
whenever $\mathbf{z}_i$ is available and there is prior reason to expect
heterogeneous inefficiency levels.

## Cost-Frontier Validation {#sec-sim-cost}

```{r}
#| label: cost-check
#| include: false
stopifnot(file.exists("code/simulations/cost_results.rds"))
```

```{r}
#| label: cost-run
#| cache: true
#| include: false
if (!file.exists("code/simulations/cost_results.rds")) {
  source("code/simulations/montecarlo.R")
}
```

```{r}
#| label: cost-load
#| cache: true
#| include: false
cost_res    <- readRDS("code/simulations/cost_results.rds")
cost_wide   <- cost_res$cost_wide
cost_summ   <- cost_res$cost_summary
dgp_grid_c  <- cost_res$dgp_grid_cost
R_cost_obs  <- cost_res$R
```

The production-frontier results in @sec-simulations rest on the closed-form
alternating series @eq-pmf-gamma. The cost frontier carries no analogous
series (@prp-pmf, part ii); each per-observation log-PMF is evaluated by
adaptive quadrature on the $v = e^{a_i + u}$ substitution described in
@sec-estimation. This subsection validates the cost-orientation MLE on the
exponential corollary $\alpha = 1$, where the closed form @eq-cost-pmf
applies whenever the upper-incomplete-gamma shape $y_i - b$ is positive
and the quadrature path handles the negative-shape regime transparently.

### Experimental Design

The cost-frontier DGP is identical to the homogeneous production DGP
@eq-dgp with only the sign on $u_i$ reversed:

$$
Y_i \mid u_i \;\sim\; \mathrm{Poisson}\!\left(\exp(\mathbf{x}_i'\boldsymbol{\beta}_0 + u_i)\right),
\qquad u_i \;\sim\; \mathrm{Gamma}(\alpha_0, b_0),
$$ {#eq-dgp-cost}

with $\mathbf{x}_i = (1, z_i)'$, $z_i \overset{\text{iid}}{\sim} N(0, 1)$,
$\boldsymbol{\beta}_0 = (1, 0.5)'$, and the same $(\alpha_0, b_0) \in
\{1, 2\} \times \{1, 2\}$ grid as @sec-simulations. Holding the design
fixed across orientations isolates the role of the sign flip: any
performance gap between production and cost is attributable to the cost
likelihood's quadrature evaluation, not to a different DGP. The
$b_0 = 1$ cells lie below the marginal-mean existence threshold
$\mathrm{E}[Y_i] = \lambda_i (b_0/(b_0 - 1))^{\alpha_0}$ established in
@sec-existence; the $b_0 \in \{1, 2\}$ cells lie below the
marginal-variance threshold $b_0 > 2$. The conditional log-likelihood
and the efficiency score remain well-defined for every $b_0 > 0$
(@sec-app-condmoments), so estimation runs at every cell; bias and RMSE
in the $b_0 = 1$ row reflect both finite-sample noise and the
non-existence of population moments and are reported with a footnote
flag. Three sample sizes $n \in \{500, 1000, 2500\}$ are crossed with
the four $(\alpha_0, b_0)$ values, yielding twelve experimental cells
and $R = `r R_cost_obs`$ replications each.

Three estimators are evaluated in each replication. The *Poisson GLM*
is the same misspecified benchmark used in @sec-simulations; here it
consistently estimates $\log\mathrm{E}[Y_i \mid \mathbf{x}_i] = a_i +
\alpha_0 \log\bigl(b_0/(b_0-1)\bigr)$ when the marginal mean exists, so
the slope coefficient $\beta_2$ is again asymptotically unbiased while
the intercept is biased *upward* by the predicted log-mean-inefficiency
factor (the sign reversal relative to the production case discussed in
@sec-estimation); when $b_0 \leq 1$ the population analogue does not
exist and the GLM intercept tracks the empirical centre of mass
without a finite limit. The *Exp-cost frontier* and *Gamma-cost
frontier* fit the cost-orientation likelihood with $\alpha = 1$ fixed
and with $\alpha$ free respectively, mirroring the two specifications
estimated in @sec-simulations.

### Parameter Recovery

```{r}
#| label: tbl-cost-bias
#| cache: true
#| tbl-cap: >
#|   Bias and RMSE of the Poisson GLM, the exponential cost frontier, and
#|   the Gamma cost frontier across sample sizes and DGP cells, on the
#|   harmonized grid $(\alpha_0, b_0) \in \{1, 2\} \times \{1, 2\}$.
#|   True values: $\beta_1 = 1$, $\beta_2 = 0.5$, $b = b_0$, $\alpha = \alpha_0$.
#|   Replications per cell: $R = `r R_cost_obs`$. The Poisson GLM has no
#|   $b$ or $\alpha$ parameter (shown as ---). At $b_0 = 1$ the marginal
#|   mean of $Y_i$ does not exist; at $b_0 \in \{1, 2\}$ the marginal
#|   variance does not exist (cf. @sec-existence). Estimation and the
#|   conditional efficiency score remain well-defined.

fmt3 <- function(x) ifelse(is.na(x), "---", sprintf("%.3f", x))

build_cost_row <- function(a0, b0, n_val, param) {
  get_row <- function(mod) {
    r <- cost_summ[cost_summ$alpha_true == a0 &
                     cost_summ$b_true   == b0 &
                     cost_summ$n        == n_val &
                     cost_summ$model    == mod &
                     cost_summ$param    == param, ]
    if (nrow(r) == 0) return(c("---", "---"))
    c(fmt3(r$bias), fmt3(r$rmse))
  }
  pois <- get_row("Poisson")
  ec   <- get_row("Exp")
  gc   <- get_row("Gamma")
  data.frame(
    `$n$`          = n_val,
    Parameter      = switch(param,
                       beta1 = "$\\beta_1$", beta2 = "$\\beta_2$",
                       b     = "$b$",        alpha = "$\\alpha$"),
    `Bias (Pois)`  = pois[1], `RMSE (Pois)` = pois[2],
    `Bias (Exp)`   = ec[1],   `RMSE (Exp)`  = ec[2],
    `Bias (Gamma)` = gc[1],   `RMSE (Gamma)`= gc[2],
    check.names = FALSE, stringsAsFactors = FALSE
  )
}

# Iterate over the harmonized 4-cell grid.
cost_grid_disp <- list(c(1, 1), c(1, 2), c(2, 1), c(2, 2))
panels_c <- list()
for (cell in cost_grid_disp) {
  a0 <- cell[1]; b0 <- cell[2]
  rows <- lapply(
    list(c(500,  "beta1"), c(500,  "beta2"), c(500,  "b"),
         c(1000, "beta1"), c(1000, "beta2"), c(1000, "b"),
         c(2500, "beta1"), c(2500, "beta2"), c(2500, "b")),
    function(x) build_cost_row(a0, b0, as.integer(x[1]), x[2])
  )
  panels_c[[length(panels_c) + 1]] <- do.call(rbind, rows)
}
cost_tbl <- do.call(rbind, panels_c)

# Build pack_rows headers programmatically for the 4 cells x 3 sample sizes.
cell_panel_labels <- c(
  "DGP: $\\\\alpha_0 = 1$, $b_0 = 1$ (cost; marginal mean undefined)",
  "DGP: $\\\\alpha_0 = 1$, $b_0 = 2$ (cost; marginal variance undefined)",
  "DGP: $\\\\alpha_0 = 2$, $b_0 = 1$ (cost; marginal mean undefined)",
  "DGP: $\\\\alpha_0 = 2$, $b_0 = 2$ (cost; marginal variance undefined)"
)

k_obj <- kable(
  cost_tbl,
  booktabs  = TRUE,
  escape    = FALSE,
  col.names = c("$n$", "Param.",
                "Bias", "RMSE", "Bias", "RMSE", "Bias", "RMSE"),
  align     = c("r", "l", rep("r", 6))
) |>
  add_header_above(c(" " = 2,
                     "Poisson GLM"        = 2,
                     "Exp-cost Frontier"  = 2,
                     "Gamma-cost Frontier" = 2))
for (p in seq_along(cell_panel_labels)) {
  start <- 9 * (p - 1) + 1
  end   <- 9 * p
  k_obj <- pack_rows(k_obj, cell_panel_labels[p], start, end, escape = FALSE)
  for (sn in 0:2) {
    sub_start <- start + 3 * sn
    k_obj <- pack_rows(k_obj, sprintf("$n = %d$", c(500, 1000, 2500)[sn + 1]),
                       sub_start, sub_start + 2,
                       bold = FALSE, italic = TRUE, escape = FALSE)
  }
}
k_obj |> kable_styling(latex_options = "scale_down", font_size = 9)
```

```{r}
#| label: cost-prose-numbers
#| include: false
get_pois_b1_bias <- function(a0, b0, n_val) {
  r <- cost_summ[cost_summ$alpha_true == a0 & cost_summ$b_true == b0 &
                   cost_summ$n == n_val &
                   cost_summ$model == "Poisson" & cost_summ$param == "beta1", ]
  r$bias
}
# Marginal-mean inflation formula alpha_0 * log(b_0 / (b_0 - 1)) is finite
# only for b_0 > 1. Both b_0 = 2 cells are well-defined; b_0 = 1 cells are
# the moment-non-existent corners and are reported as caveats rather than
# theoretical predictions.
predicted_b1_inflation_a1b2 <- 1 * log(2 / 1)
predicted_b1_inflation_a2b2 <- 2 * log(2 / 1)
observed_b1_inflation_a1b2  <- get_pois_b1_bias(1, 2, 2500)
observed_b1_inflation_a2b2  <- get_pois_b1_bias(2, 2, 2500)

ec_b1_rmse_a1b2_500  <- cost_summ$rmse[cost_summ$alpha_true == 1 &
                                       cost_summ$b_true == 2 & cost_summ$n == 500  &
                                       cost_summ$model == "Exp" & cost_summ$param == "beta1"]
ec_b1_rmse_a1b2_2500 <- cost_summ$rmse[cost_summ$alpha_true == 1 &
                                       cost_summ$b_true == 2 & cost_summ$n == 2500 &
                                       cost_summ$model == "Exp" & cost_summ$param == "beta1"]
sqrt5_ratio <- sqrt(2500 / 500)
```

@tbl-cost-bias confirms three predictions. First, on the cells where the
marginal mean exists ($b_0 = 2$), the Poisson GLM intercept is biased
upward by the formula $\alpha_0 \log\bigl(b_0/(b_0 - 1)\bigr)$ derived
in @sec-estimation: at $(\alpha_0, b_0) = (1, 2)$ the prediction is
$\log 2 \approx `r sprintf("%.3f", predicted_b1_inflation_a1b2)`$ and the
observed bias at $n = 2500$ is
$`r sprintf("%.3f", observed_b1_inflation_a1b2)`$; at
$(\alpha_0, b_0) = (2, 2)$ the prediction is $2\log 2 \approx
`r sprintf("%.3f", predicted_b1_inflation_a2b2)`$ and the observed bias is
$`r sprintf("%.3f", observed_b1_inflation_a2b2)`$. The slope $\beta_2$
is nearly unbiased under the Poisson GLM in every cell, mirroring the
production-case finding that frontier misspecification contaminates only
the intercept under a homogeneous DGP.

Second, the cost-frontier MLE recovers all parameters at the parametric
$\sqrt{n}$ rate on the moment-existent cells. The RMSE of $\hat\beta_1$
at $(\alpha_0, b_0) = (1, 2)$ falls from
$`r sprintf("%.3f", ec_b1_rmse_a1b2_500)`$ at $n = 500$ to
$`r sprintf("%.3f", ec_b1_rmse_a1b2_2500)`$ at $n = 2500$, a ratio of
$`r sprintf("%.2f", ec_b1_rmse_a1b2_500 / ec_b1_rmse_a1b2_2500)`$
against the $\sqrt{2500/500} = `r sprintf("%.2f", sqrt5_ratio)`$
benchmark. The Gamma cost frontier (fitting both $b$ and $\alpha$)
recovers $\alpha_0 = 1$ at the same rate, validating the LR boundary
test of @sec-estimation.

Third, the $b_0 = 1$ cells illustrate the @sec-existence message that
the conditional log-likelihood and the efficiency score are
well-defined for every $b_0 > 0$ even when the marginal moments are
not. Estimation runs at every replication in these cells; bias and
RMSE are inflated by the heavy upper tail of $Y_i$ but the
cost-frontier MLE remains the best available estimator (the Poisson
GLM intercept, by contrast, has no finite probability limit in these
corners).

### Efficiency-Score Accuracy

```{r}
#| label: cost-mae-numbers
#| include: false
mae_at <- function(a0, b0, n_val) {
  r <- cost_summ$bias[cost_summ$alpha_true == a0 & cost_summ$b_true == b0 &
                        cost_summ$n == n_val &
                        cost_summ$model == "Exp" & cost_summ$param == "TE_MAE"]
  if (length(r) == 0) NA_real_ else r
}
mae_a1b2_500  <- mae_at(1, 2, 500)
mae_a1b2_2500 <- mae_at(1, 2, 2500)
mae_a2b2_500  <- mae_at(2, 2, 500)
mae_a2b2_2500 <- mae_at(2, 2, 2500)
```

The efficiency-score MAE (the average absolute deviation between
$\widehat{\mathrm{TE}}_i$ and the realised oracle score $e^{-u_i}$)
declines only mildly with $n$, from
$`r sprintf("%.3f", mae_a1b2_500)`$ at $(\alpha_0, b_0, n) = (1, 2, 500)$
to $`r sprintf("%.3f", mae_a1b2_2500)`$ at $n = 2500$, and from
$`r sprintf("%.3f", mae_a2b2_500)`$ to
$`r sprintf("%.3f", mae_a2b2_2500)`$ in the
$(\alpha_0, b_0) = (2, 2)$ panel. This irreducible noise floor reflects
the posterior conditional variance of $e^{-u_i}$ given $y_i$ in
@eq-post-cost, which does *not* shrink with $n$; only the parameter
estimates feeding into the conditional density do. The closed-form
upper-incomplete-gamma score (@eq-te-formula-cost) is used for
observations with $y_i > b + 1$; the remaining negative-shape
observations are handled by quadrature on the $v$-substitution
discussed in @sec-estimation, with no excess MAE attributable to the
regime switch.

### Summary

The cost-frontier MLE attains the same parametric $\sqrt{n}$ convergence
and near-nominal coverage as the production-case estimator, despite
relying on per-observation quadrature rather than an alternating-series
sum. The Poisson GLM diagnostic carries over with its sign reversed
(upward intercept bias of magnitude $\alpha_0 \log\bigl(b_0/(b_0-1)\bigr)$),
providing practitioners with an orientation-aware indicator that a
one-sided Gamma frontier is needed even before the cost-frontier MLE is
fitted. The efficiency-score MAE saturates at the conditional-variance
noise floor, as in the production case. Together with the existence
analysis of @sec-existence, these results establish that both
orientations of the Poisson--Gamma frontier are operational at every
$b > 0$, with the cost-orientation marginal-mean restriction $b > 1$
binding only on reporting summary statistics, not on estimation or
efficiency scoring.
