# Empirical Application {#sec-empirical}

```{r}
#| label: setup-06
#| include: false
#| cache: false
library(countSFA)
library(ggplot2)
library(kableExtra)
options(knitr.table.format = if (knitr::is_latex_output()) "latex" else "html")
```

```{r}
#| label: emp-mle-fit
#| cache: true
#| include: false
#| error: true
if (!file.exists("code/empirical/fits.rds")) {
  source("code/empirical/application.R")
}
```

```{r}
#| label: emp-load
#| include: false
emp <- readRDS("code/empirical/fits.rds")
dat         <- emp$data
y           <- emp$y
X           <- emp$X
fit_pois    <- emp$fit_pois
fit_exp     <- emp$fit_exp
fit_gam     <- emp$fit_gam
scores_exp  <- emp$scores_exp
scores_gam  <- emp$scores_gam
lr_stat     <- emp$lr_stat
lr_pval     <- emp$lr_pval
compare_tbl <- emp$compare_tbl

# Preferred model: Gamma if converged to a finite, non-degenerate optimum
# AND its scores were computed; otherwise fall back to the exponential
# fit. The multi-start may report convergence = 0 while sitting on the
# alpha -> 0 ridge with loglik = -.Machine$double.xmax (a sentinel from
# the PMF evaluator); reject that case explicitly.
gam_converged <- !is.null(fit_gam) &&
                 isTRUE(fit_gam$convergence == 0L) &&
                 is.finite(fit_gam$loglik) &&
                 fit_gam$loglik > -1e10 &&
                 !is.null(scores_gam)
exp_converged    <- !is.null(fit_exp) && fit_exp$convergence == 0L
preferred_scores <- if (gam_converged) scores_gam else scores_exp
preferred_label  <- if (gam_converged) "Gamma"    else "exponential"
lr_reject        <- !is.na(lr_pval) && lr_pval < 0.05

# Pre-compute figure captions so !expr is not needed (avoids YAML quoting issues)
cap_eff_hist <- paste0(
  "Distribution of estimated technical efficiency scores from the Poisson-",
  preferred_label," frontier. Each of the ", nrow(dat), " observations is one firm-year pair.",
  " The dashed vertical line marks the sample mean."
)
cap_eff_scatter <- paste0(
  "Technical efficiency versus log R&D expenditure, by industry sector.",
  " Points are firm-year observations; LOESS smooth (span = 0.75) per sector.",
  " Model: Poisson-", preferred_label, " frontier."
)
```

## Data {#sec-emp-data}

We apply the Poisson stochastic frontier to the patent panel introduced by
@hausman1984, which has served as the canonical benchmark for count-data
econometrics for four decades [@cameronTrivedi2005, Ch. 23;
@hallGrilichesHausman1986]. The sample comprises
`r nrow(dat) / length(unique(dat$year))` U.S. manufacturing firms observed
annually over five years (1968–1972), yielding `r nrow(dat)` pooled firm-year
observations. We treat the data as a pooled cross-section and include year
fixed effects to absorb aggregate time trends, following the approach of
@feHofler2013 for count-data frontiers.

The outcome is the annual patent count $Y_{it}$. As @tbl-data-summary
shows, the distribution is markedly right-skewed, ranging from
`r min(y)` to `r max(y)` with a mean of `r round(mean(y), 2)` and standard
deviation of `r round(sd(y), 2)`. Zero-patent firm-years account for
`r round(100 * mean(y == 0), 1)`% of the sample, confirming the degree of
zero-inflation documented by @hallGrilichesHausman1986. The variance-to-mean
ratio of `r round(var(y)/mean(y), 1)` far exceeds 1, indicating severe
overdispersion relative to a Poisson model: a pattern the frontier structure
attributes to heterogeneous technical inefficiency rather than pure
unobserved heterogeneity.

The frontier log-mean is parameterised as

$$
a_{it} \;=\; \mathbf{x}_{it}'\boldsymbol{\beta}
\;=\; \beta_1 + \beta_2 \log R_{it} + \beta_3 \log K_{it}
     + \beta_4 \cdot \mathbf{1}[\text{Science}]_i + \sum_{t=2}^{5}\delta_t d_t,
$$ {#eq-frontier-spec}

where $\log R_{it}$ is log R&D expenditure (the primary innovation input),
$\log K_{it}$ is log capital stock, $\mathbf{1}[\text{Science}]_i$ is a
time-invariant science-sector indicator, and $d_t$ are year dummies.^[The
year indicator coefficients in @eq-frontier-spec are components of the frontier
coefficient vector $\boldsymbol{\beta}$; the symbol $\delta$ is reserved
exclusively for the scaling parameters in @eq-scaling-bi.]

```{r}
#| label: tbl-data-summary
#| tbl-cap: "Summary statistics for the Hausman--Hall--Griliches patent panel (1730 firm-year observations, 346 firms, 1968--1972). The variance-to-mean ratio of the patent count far exceeds 1, indicating overdispersion that the standard Poisson regression cannot accommodate."

vars <- data.frame(
  Variable = c("Patent count ($Y$)", "Log R\\&D ($\\log R$)",
               "Log capital ($\\log K$)", "Science sector (\\%)"),
  Mean  = c(mean(y),             mean(dat$rd), mean(dat$lk), 100 * mean(dat$sci)),
  SD    = c(sd(y),               sd(dat$rd),   sd(dat$lk),   100 * sd(dat$sci)),
  Min   = c(min(y),              min(dat$rd),  min(dat$lk),  0),
  Max   = c(max(y),              max(dat$rd),  max(dat$lk),  100),
  stringsAsFactors = FALSE
)
vars[, 2:5] <- lapply(vars[, 2:5], round, digits = 2)

knitr::kable(
  vars,
  booktabs  = TRUE,
  escape    = FALSE,
  col.names = c("Variable", "Mean", "SD", "Min", "Max")
) |>
  kableExtra::kable_styling() |>
  kableExtra::footnote(
    general       = paste0("Variance-to-mean ratio of patent count: ",
                           round(var(y) / mean(y), 1), "."),
    general_title = "",
    footnote_as_chunk = TRUE
  )
```

## MLE Results {#sec-emp-results}

@tbl-results-main reports maximum likelihood estimates for two specifications:
the standard Poisson GLM (misspecified benchmark) and the Poisson frontier
with exponential inefficiency ($\alpha = 1$ fixed). The Gamma frontier was
attempted but did not achieve numerical convergence on this dataset, a finding
discussed in @sec-conclusion. All models are
estimated on the pooled sample; standard errors are in parentheses.

```{r}
#| label: tbl-results-main
#| tbl-cap: "MLE estimates for the homogeneous specifications on the patent panel: standard Poisson GLM (misspecified benchmark) and Poisson frontier with exponential inefficiency. Standard errors (delta method for $\\hat{b}$) in parentheses. Year fixed effects included but not reported. Log-lik.\\ and AIC computed at the maximum. The Gamma frontier with free $\\alpha$ did not converge on the pooled cross-section; convergence behaviour is discussed in @sec-emp-scaling."

make_coef_row <- function(fit_obj, is_glm = FALSE) {
  fmt <- function(v, s) sprintf("%.3f (%.3f)", v, s)
  if (is_glm) {
    cf <- coef(fit_obj)
    se <- sqrt(diag(vcov(fit_obj)))
    list(
      Intercept = fmt(cf[1],       se[1]),
      log_RD    = fmt(cf["rd"],    se["rd"]),
      log_K     = fmt(cf["lk"],    se["lk"]),
      sci_sect  = fmt(cf["sci"],   se["sci"]),
      b         = "---",
      alpha     = "---",
      LogLik    = round(as.numeric(logLik(fit_obj)), 1),
      AIC       = round(AIC(fit_obj), 1)
    )
  } else {
    cf <- fit_obj$coefficients
    se <- fit_obj$se
    list(
      Intercept = fmt(cf[1], se[1]),
      log_RD    = fmt(cf[2], se[2]),
      log_K     = fmt(cf[3], se[3]),
      sci_sect  = fmt(cf[4], se[4]),
      b         = fmt(fit_obj$b, fit_obj$se_b),
      alpha     = if (!is.na(fit_obj$alpha) && fit_obj$dist == "gamma")
                    fmt(fit_obj$alpha, fit_obj$se_alpha)
                  else "1 (fixed)",
      LogLik    = round(fit_obj$loglik, 1),
      AIC       = round(fit_obj$AIC,   1)
    )
  }
}

r_pois <- make_coef_row(fit_pois, is_glm = TRUE)
r_exp  <- if (exp_converged) make_coef_row(fit_exp) else NULL
r_gam  <- if (gam_converged) make_coef_row(fit_gam) else NULL

rows <- Filter(Negate(is.null),
               list("Poisson GLM"    = r_pois,
                    "Exp frontier"   = r_exp,
                    "Gamma frontier" = r_gam))

tbl_coef <- do.call(rbind, lapply(names(rows), function(nm) {
  r <- rows[[nm]]
  data.frame(
    Model     = nm,
    Intercept = r$Intercept,
    `log R`   = r$log_RD,
    `log K`   = r$log_K,
    Science   = r$sci_sect,
    b         = r$b,
    alpha     = r$alpha,
    LogLik    = r$LogLik,
    AIC       = r$AIC,
    check.names = FALSE, stringsAsFactors = FALSE
  )
}))

knitr::kable(
  tbl_coef,
  booktabs  = TRUE,
  escape    = FALSE,
  col.names = c("Model", "Intercept", "$\\log R$", "$\\log K$",
                "Science", "$\\hat{b}$", "$\\hat{\\alpha}$",
                "Log-lik.", "AIC")
) |>
  kableExtra::kable_styling(latex_options = "scale_down") |>
  kableExtra::footnote(
    general = "Year fixed effects included but not reported. Delta-method SEs for $\\\\hat{b}$ and $\\\\hat{\\\\alpha}$.",
    general_title = "",
    footnote_as_chunk = TRUE,
    escape = FALSE
  )
```

```{r}
#| label: emp-glm-bias
#| include: false
# Expected Poisson intercept attenuation for comparison in prose
pois_intercept <- round(coef(fit_pois)[1], 3)
exp_intercept  <- if (exp_converged) round(fit_exp$coefficients[1], 3) else NA
```

The pattern in @tbl-results-main aligns with the theoretical predictions from
@sec-simulations. The Poisson GLM intercept (`r pois_intercept`) is attenuated
relative to the frontier specification (`r exp_intercept`), reflecting the
log-mean-inefficiency bias identified in the Monte Carlo study. The slope on
log R&D is positive and significant across both specifications, replicating
the long-established empirical relationship between research spending and
patent counts [@pakesGriliches1980; @hausman1984].^[As an external check, refitting the pooled Poisson GLM with the current value and four lags of $\log R$ recovers a sum-of-elasticities of $0.486$, which exactly matches the corresponding entry in @cameronTrivedi2005, Table 23.1. Adding the same four lags to the frontier specifications, however, induces severe multicollinearity along the $b$-direction and prevents both the exponential and Gamma likelihoods from converging to a finite interior maximum.] The capital stock coefficient is positive but smaller in
magnitude than the R&D coefficient, consistent with the labour-intensive nature
of knowledge production. Science-sector firms have a systematically higher
frontier mean, reflecting the superior technological opportunities available in
chemistry, electronics, and scientific instruments.

The estimated rate $\hat{b}$ from the exponential frontier implies an average
inefficiency of $\mathrm{E}[u_i] = 1/\hat{b}$ in log-output units: the
average firm's expected output is $\exp(-1/\hat{b})$ times its frontier
output, with the full distribution of inefficiency heterogeneity described
by the fitted exponential density.

::: {.remark title="Panel structure and standard errors."}
The patent dataset is a balanced panel of 346 firms observed over five years. The standard errors reported in
@tbl-results-main are based on the inverse Hessian and do not account for within-firm serial correlation. Cluster-robust standard errors (clustered at
the firm level) would be more conservative; their computation is deferred to
future work. Coefficient estimates are unaffected by this choice.
:::

::: {.remark title="Gamma frontier convergence"}
The Gamma frontier with free $\alpha$ did not converge on the patent panel.
The likelihood surface is flat along the ridge $\alpha \to 0$, $b \to \infty$
at constant $\alpha/b$ (see @sec-estimation), and with $n = 1730$ pooled
observations the curvature in the $\alpha$ direction is insufficient for
the optimiser to locate an interior maximum reliably. We therefore retain
the exponential frontier as the preferred homogeneous specification here. A
related diagnostic from the scaling-model fit of @sec-emp-scaling (where the
free-$\alpha$ Gamma frontier does converge but to $\hat\alpha = 1.000$ at the
exponential boundary) is reported alongside the LR test in that section.
:::

## Efficiency Scores {#sec-emp-efficiency}

The technical efficiency of firm $i$ in year $t$ is the conditional
expectation

$$
\widehat{\mathrm{TE}}_{it}
\;=\; \mathrm{E}\!\left[e^{-u_{it}} \mid Y_{it} = y_{it},\, \mathbf{x}_{it}\right],
$$ {#eq-te-emp}

computed from the preferred (`r preferred_label`) specification via the
formula developed in @sec-model. The mean estimated efficiency
across the sample is `r round(mean(preferred_scores$eff_score), 3)`, indicating
that the average firm achieves roughly
`r round(100 * mean(preferred_scores$eff_score), 1)`% of its frontier patent
output, as shown in @fig-eff-hist. The distribution is left-skewed, with most
firms bunched near the frontier and a tail of chronically inefficient units.

:::{#fig-eff-hist}
```{r}

ggplot(preferred_scores, aes(x = eff_score)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white",
                 linewidth = 0.3) +
  geom_vline(xintercept = mean(preferred_scores$eff_score),
             linetype = "dashed", colour = "firebrick", linewidth = 0.8) +
  annotate("text",
           x     = mean(preferred_scores$eff_score) + 0.013,
           y     = Inf,
           label = paste0("Mean = ", round(mean(preferred_scores$eff_score), 3)),
           hjust = 0, vjust = 1.6, size = 3.2, colour = "firebrick") +
  labs(x = "Technical efficiency score", y = "Count") +
  theme_bw(base_size = 11)
```

Distribution of estimated technical efficiency scores from the Poisson-`r preferred_label` frontier. Each of the `r nrow(dat)` observations is one firm-year pair. The dashed vertical line marks the sample mean.
:::

The efficiency distribution in @fig-eff-hist displays the right-skewed shape
characteristic of an exponential inefficiency density: the exponential distribution
assigns highest density near $u = 0$, so the distribution places substantial
probability on efficiency scores close to 1 while maintaining a long lower tail
for chronically inefficient firms. The interquartile range of
`r round(diff(quantile(preferred_scores$eff_score, c(0.25, 0.75))), 3)` reflects
meaningful cross-firm heterogeneity in patenting productivity.

```{r}
#| label: fig-eff-scatter
#| fig-cap: !expr cap_eff_scatter

plot_df <- data.frame(
  eff    = preferred_scores$eff_score,
  rd     = dat$rd,
  sector = factor(ifelse(dat$sci == 1L, "Science", "Other"),
                  levels = c("Science", "Other"))
)

ggplot(plot_df, aes(x = rd, y = eff, colour = sector)) +
  geom_point(alpha = 0.30, size = 0.8) +
  geom_smooth(method = "loess", formula = y ~ x,
              span = 0.75, se = FALSE, linewidth = 1.0) +
  scale_colour_manual(values = c("Science" = "steelblue", "Other" = "grey40"),
                      name   = NULL) +
  labs(x = "Log R\u0026D expenditure",
       y = expression(widehat(TE)[it])) +
  theme_bw(base_size = 11) +
  theme(legend.position="bottom")
```

```{r}
#| label: fig-eff-year
#| fig-cap: "Conditional technical efficiency by year (1968--1972). The box spans the interquartile range; the horizontal line is the median; whiskers extend to $1.5 \\times$ IQR. Overlaid points (jittered) show the full distribution. A slight upward drift in median efficiency is consistent with aggregate technological progress over the period."

year_df <- data.frame(
  eff  = preferred_scores$eff_score,
  year = factor(dat$year, labels = paste0(1967 + 1:5))
)

ggplot(year_df, aes(x = year, y = eff)) +
  geom_jitter(width = 0.15, alpha = 0.18, size = 0.7, colour = "steelblue") +
  geom_boxplot(fill = NA, colour = "grey30", outlier.shape = NA,
               linewidth = 0.6, width = 0.45) +
  labs(x = "Year", y = expression(widehat(TE)[it])) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  theme_bw(base_size = 11)
```

@fig-eff-scatter reveals a positive cross-sectional association between R&D
intensity and estimated efficiency: firms that invest more in research tend to
operate closer to the patent frontier, controlling for capital stock and sector.
Science-sector firms cluster in the upper-right quadrant, consistent with the
view that technological opportunity amplifies the returns to R&D investment in
high-opportunity industries. The association is approximately log-linear in both
sectors, with no evidence of diminishing returns to R&D intensity within the
observed range.

@fig-eff-year documents the time profile of efficiency over the 1968–1972
period. The median efficiency is broadly stable across years, suggesting that
aggregate technological change is adequately absorbed by the year fixed effects
in the frontier specification @eq-frontier-spec. Any residual upward drift
in the distribution over time is consistent with learning-by-doing or the gradual
adoption of more productive research practices over the period.

::: {.remark title="Panel structure and unit effects"}
The present application treats
the data as a pooled cross-section, so the efficiency score $\widehat{\mathrm{TE}}_{it}$
varies both across firms and across years within a firm. A natural extension would
be to introduce firm fixed effects into the frontier mean $a_{it}$, which would
partial out time-invariant heterogeneity and allow $u_{it}$ to capture only
within-firm year-to-year inefficiency. Such a panel frontier is left for future
work (see @sec-conclusion for discussion), but the stability visible in
@fig-eff-year suggests that the pooled estimates are not severely contaminated
by time-invariant unobserved heterogeneity over this short five-year window.
:::

The empirical results confirm that the exponential frontier adds meaningful
structure over the standard Poisson GLM: the frontier intercept shift is
consistent with the Monte Carlo predictions, the slope estimates are stable
across both specifications, and the resulting efficiency scores reveal
economically interpretable cross-firm and cross-sector variation. The
scaling model extension in @sec-emp-scaling examines how firm-level capital
intensity and sector membership shape the distribution of inefficiency.
@sec-conclusion collects the paper's contributions and outlines extensions
to panel data, zero-inflated counts, and Bayesian estimation.

## Inefficiency Determinants {#sec-emp-scaling}

The homogeneous frontier estimated in @sec-emp-results constrains every firm
to draw from the same inefficiency distribution, implying a common mean
$\mathrm{E}(u_i) = 1/b$. Systematic firm-level characteristics (in particular
capital intensity and sector membership) may shift the level of inefficiency
independently of the frontier mean $a_i$. This section extends the analysis
using the scaling model of @sec-scaling, which writes
$b_i = b\exp(-z_i'\delta)$ so that $\mathrm{E}(u_i) = \exp(z_i'\delta)/b$
is firm-specific [@caudillFord1993; @wangSchmidt2002].

Two inefficiency determinants are selected on substantive economic grounds.
The first is log capital stock ($z_1 = \log K_i$). Capital-intensive firms
may be better positioned to translate R&D investment into patents through
complementarity between physical and knowledge capital: laboratory equipment,
manufacturing prototypes, and testing infrastructure all lower the per-patent
cost of the innovation process. This implies $\delta_1 < 0$, so that more
capital reduces inefficiency by raising $b_i$. The second determinant is the
science-sector indicator ($z_2 = \mathbf{1}[\text{Science}]_i$). Firms in
chemistry, electronics, and scientific instruments maintain dedicated IP
practices, routinised patenting workflows, and deeper absorptive capacity
[@hallGrilichesHausman1986], all of which reduce the gap between actual
output and the patent frontier. The sign prediction is again $\delta_2 < 0$.
Log R&D expenditure is deliberately excluded from $z_i$: it already enters
the frontier mean $a_i$ through $\mathbf{x}_i$, so including it in $z_i$
would conflate $\beta_{\log R}$ and $\delta$. An optional quadratic term
$z_3 = (\log K_i)^2$ is retained in the extended model (M5) only if the
likelihood ratio test against M4 rejects at the 5\% level.

```{r}
#| label: emp-scaling-load
#| include: false
stopifnot(file.exists("results/empirical_fits.rds"),
          file.exists("results/efficiency_scores.rds"))

ef  <- readRDS("results/empirical_fits.rds")
sc  <- readRDS("results/efficiency_scores.rds")

dat_sc <- ef$data
y_sc   <- ef$y
X_sc   <- ef$X
p_sc   <- ncol(X_sc)

fit_pois_sc <- ef$fit_pois
fit_M2_sc   <- ef$fit_M2
fit_M3_sc   <- ef$fit_exp
fit_M4_sc   <- ef$fit_M4
fit_M5_sc   <- ef$fit_M5
fit_M6_sc   <- ef$fit_M6

ll_M1_sc <- ef$ll_M1
ll_M2_sc <- ef$ll_M2
ll_M3_sc <- ef$ll_M3
ll_M4_sc <- ef$ll_M4
ll_M5_sc <- ef$ll_M5

se_M4_sc    <- ef$se_M4
se_M5_sc    <- ef$se_M5
lr_tests_sc <- ef$lr_tests
pref_sc     <- ef$preferred_model

# Preferred scaling fit
fit_pref_sc <- if (pref_sc == "M5" && !is.null(fit_M5_sc)) fit_M5_sc else fit_M4_sc
ll_pref_sc  <- if (pref_sc == "M5") ll_M5_sc  else ll_M4_sc
se_pref_sc  <- if (pref_sc == "M5") se_M5_sc  else se_M4_sc
Z_pref_sc   <- if (pref_sc == "M5") ef$Z_full else ef$Z_base
q_pref_sc   <- ncol(Z_pref_sc)

m6_converged_sc <- !is.null(fit_M6_sc)

# Column positions in X_sc (always: 1=Intercept, 2=rd, 3=lk, 4=sci, 5+=year FEs)
# Using fixed positions because X colnames differ across GLM and frontier objects
idx_int <- 1L
idx_rd  <- 2L
idx_lk  <- 3L
idx_sci <- 4L

# Helper: format estimate (SE) with significance stars
fmt_s <- function(est, se) {
  if (is.null(est) || is.na(est) || is.null(se) || is.na(se)) return("---")
  z  <- abs(est / se)
  pv <- 2 * pnorm(-z)
  s  <- if (pv < 0.01) "***" else if (pv < 0.05) "**" else if (pv < 0.10) "*" else ""
  paste0(sprintf("%.3f", est), " (", sprintf("%.3f", se), ")", s)
}

# GLM
cf_glm <- coef(fit_pois_sc)
se_glm <- sqrt(diag(vcov(fit_pois_sc)))

# M2 (PHN) SEs from Hessian
if (!is.null(fit_M2_sc) && !is.null(fit_M2_sc$hessian)) {
  se_M2_all <- tryCatch(
    sqrt(pmax(diag(solve(fit_M2_sc$hessian)), 0)),
    error = function(e) rep(NA_real_, length(fit_M2_sc$par))
  )
} else {
  se_M2_all <- rep(NA_real_, p_sc + 1L)
}

# M3 (Exp homog): fit_poisson_frontier S3 object
cf_M3 <- if (!is.null(fit_M3_sc)) fit_M3_sc$coefficients else rep(NA_real_, p_sc)
se_M3 <- if (!is.null(fit_M3_sc)) fit_M3_sc$se           else rep(NA_real_, p_sc)
b_M3  <- if (!is.null(fit_M3_sc)) fit_M3_sc$b            else NA_real_
sb_M3 <- if (!is.null(fit_M3_sc)) fit_M3_sc$se_b         else NA_real_

# Preferred scaling: b and delta
b_pref_sc     <- exp(fit_pref_sc$par[p_sc + 1L])
sb_pref_sc    <- se_pref_sc[p_sc + 1L] * b_pref_sc   # delta method
delta_pref    <- fit_pref_sc$par[(p_sc + 2L):(p_sc + 1L + q_pref_sc)]
se_delta_pref <- se_pref_sc[(p_sc + 2L):(p_sc + 1L + q_pref_sc)]

# AIC
n_obs  <- length(y_sc)
aic_M1 <- AIC(fit_pois_sc)
aic_M2 <- if (!is.null(fit_M2_sc) && is.finite(ll_M2_sc))
  2 * (p_sc + 1) - 2 * ll_M2_sc else NA_real_
aic_M3 <- if (!is.null(fit_M3_sc)) fit_M3_sc$AIC else NA_real_
aic_M4 <- if (!is.null(fit_M4_sc) && is.finite(ll_M4_sc))
  2 * (p_sc + 1 + 2) - 2 * ll_M4_sc else NA_real_
aic_M5 <- if (!is.null(fit_M5_sc) && is.finite(ll_M5_sc))
  2 * (p_sc + 1 + 3) - 2 * ll_M5_sc else NA_real_
aic_pref <- if (pref_sc == "M5") aic_M5 else aic_M4

# Mean efficiency
te_mean_M3   <- round(mean(sc$te_homog,   na.rm = TRUE), 3)
te_mean_pref <- round(mean(sc$te_scaling, na.rm = TRUE), 3)

# Delta summary for prose
delta1_hat  <- delta_pref[1L]
delta2_hat  <- delta_pref[2L]
delta1_se   <- se_delta_pref[1L]
delta2_se   <- se_delta_pref[2L]

mean_eu    <- mean(1 / sc$b_i_pref, na.rm = TRUE)   # mean E[u_i]
me_logk    <- delta1_hat * mean_eu                    # marginal effect at mean z
ineff_ratio <- exp(delta2_hat)                        # sci vs non-sci inefficiency ratio

# Model-comparison test objects
ic31 <- lr_tests_sc$IC_31   # AIC/BIC gap: M3 vs M1 (boundary; no formal test)
v32  <- lr_tests_sc$V_32    # Vuong: M3 vs M2 (genuinely non-nested)
lr2  <- lr_tests_sc$LR_2    # Nested LR: M4 vs M3
lr3  <- lr_tests_sc$LR_3    # Nested LR: M5 vs M4
lr4  <- lr_tests_sc$LR_4    # Nested LR: M6 vs M3 (Gamma vs Exp; only present if M6 converged)
```

@tbl-model-comparison collects maximum likelihood estimates for all fitted
specifications. Moving from the Poisson GLM (M1) to the homogeneous
exponential frontier (M3) raises the log-likelihood by
`r round(ll_M3_sc - ll_M1_sc, 1)` log-units at the cost of one additional
parameter, confirming that a one-sided inefficiency term is warranted by
the data. The Poisson half-normal benchmark (M2) of @feHofler2013 achieves a
similar fit to M3. Adding inefficiency determinants in the scaling
specification further raises the log-likelihood while shifting the intercept
and the baseline rate $b$, since firm-specific mean inefficiency is now
partitioned between the homogeneous component and $\exp(-z_i'\delta)$.

```{r}
#| label: tbl-model-comparison
#| tbl-cap: "MLE estimates for the patent frontier. Standard errors in parentheses; $^{***}p<0.01$, $^{**}p<0.05$, $^{*}p<0.10$. Year fixed effects included but not reported. M2~=~Poisson half-normal (Fé~\\& Hofler 2013 benchmark); M3~=~exponential homogeneous; M4~=~exponential scaling ($z_1, z_2$); M5~=~exponential scaling ($z_1, z_2, z_3$)."

pref_label <- paste0(pref_sc, " (preferred)")

row_names <- c(
  "Intercept", "$\\log R$", "$\\log K$", "Science",
  "$\\hat{b}$", "$\\hat{\\sigma}$ (PHN only)",
  "$\\hat{\\delta}_1$ (logk)",
  "$\\hat{\\delta}_2$ (sci)",
  "$\\hat{\\delta}_3$ (logk$^2$)",
  "Log-lik.", "AIC", "$N$", "Mean $\\widehat{\\mathrm{TE}}$"
)

# M1
m1 <- c(
  fmt_s(cf_glm[idx_int], se_glm[idx_int]),
  fmt_s(cf_glm[idx_rd],  se_glm[idx_rd]),
  fmt_s(cf_glm[idx_lk],  se_glm[idx_lk]),
  fmt_s(cf_glm[idx_sci], se_glm[idx_sci]),
  "---", "---", "---", "---", "---",
  sprintf("%.1f", ll_M1_sc), sprintf("%.1f", aic_M1),
  as.character(n_obs), "---"
)

# M2
if (!is.null(fit_M2_sc) && is.finite(ll_M2_sc)) {
  sigma_M2 <- exp(fit_M2_sc$par[p_sc + 1L])
  se_sig_M2 <- se_M2_all[p_sc + 1L] * sigma_M2
  m2 <- c(
    fmt_s(fit_M2_sc$par[idx_int], se_M2_all[idx_int]),
    fmt_s(fit_M2_sc$par[idx_rd],  se_M2_all[idx_rd]),
    fmt_s(fit_M2_sc$par[idx_lk],  se_M2_all[idx_lk]),
    fmt_s(fit_M2_sc$par[idx_sci], se_M2_all[idx_sci]),
    "---",
    fmt_s(sigma_M2, se_sig_M2),
    "---", "---", "---",
    sprintf("%.1f", ll_M2_sc), sprintf("%.1f", aic_M2),
    as.character(n_obs), "---"
  )
} else {
  m2 <- rep("---", length(row_names))
}

# M3
if (!is.null(fit_M3_sc)) {
  m3 <- c(
    fmt_s(cf_M3[idx_int], se_M3[idx_int]),
    fmt_s(cf_M3[idx_rd],  se_M3[idx_rd]),
    fmt_s(cf_M3[idx_lk],  se_M3[idx_lk]),
    fmt_s(cf_M3[idx_sci], se_M3[idx_sci]),
    fmt_s(b_M3, sb_M3),
    "---", "---", "---", "---",
    sprintf("%.1f", ll_M3_sc), sprintf("%.1f", aic_M3),
    as.character(n_obs), sprintf("%.3f", te_mean_M3)
  )
} else {
  m3 <- rep("---", length(row_names))
}

# Preferred scaling (M4 or M5)
m_pref <- c(
  fmt_s(fit_pref_sc$par[idx_int], se_pref_sc[idx_int]),
  fmt_s(fit_pref_sc$par[idx_rd],  se_pref_sc[idx_rd]),
  fmt_s(fit_pref_sc$par[idx_lk],  se_pref_sc[idx_lk]),
  fmt_s(fit_pref_sc$par[idx_sci], se_pref_sc[idx_sci]),
  fmt_s(b_pref_sc, sb_pref_sc),
  "---",
  fmt_s(delta_pref[1L], se_delta_pref[1L]),
  fmt_s(delta_pref[2L], se_delta_pref[2L]),
  if (q_pref_sc == 3L) fmt_s(delta_pref[3L], se_delta_pref[3L]) else "---",
  sprintf("%.1f", ll_pref_sc), sprintf("%.1f", aic_pref),
  as.character(n_obs), sprintf("%.3f", te_mean_pref)
)

tbl_mc <- data.frame(
  Statistic = row_names,
  M1        = m1,
  M2        = m2,
  M3        = m3,
  Preferred = m_pref,
  check.names = FALSE, stringsAsFactors = FALSE
)
names(tbl_mc)[5] <- pref_label

knitr::kable(
  tbl_mc, booktabs = TRUE, escape = FALSE,
  col.names = c("", "M1: Poisson", "M2: PHN", "M3: Exp hom.", pref_label)
) |>
  kableExtra::kable_styling(latex_options = "scale_down") |>
  kableExtra::footnote(
    general = paste0(
      "Year FEs (4 dummies) not reported. ",
      "Delta-method SEs for $\\\\hat{b}$, $\\\\hat{\\\\sigma}$, and $\\\\hat{\\\\delta}$. ",
      "AIC $= -2\\\\ell + 2k$."),
    general_title = "", footnote_as_chunk = TRUE, escape = FALSE
  )
```

@tbl-model-tests reports the model-comparison statistics. We use a
test only where the asymptotic theory it relies on is valid for the
specific comparison.

  - The chi-squared **likelihood-ratio test** is reported for LR$_2$
    (M4 vs M3) and LR$_3$ (M5 vs M4). These are nested within the
    exponential scaling family and the restricted parameter values
    ($\delta = 0$ and $\delta_3 = 0$ respectively) lie in the interior
    of the alternative space, so the standard $\chi^2$ reference
    distribution holds.
  - The **Vuong [-@vuong1989] test** is reported for V$_{32}$
    (M3 vs M2): the exponential Gamma frontier and the Poisson
    half-normal benchmark of @feHofler2013 are distinct distributional
    families, both fitted at parameter values well inside their
    respective interiors, so the standard normal limit
    $V = \sqrt{n}\,\bar m / s_m \to_d \mathcal{N}(0,1)$ under the null
    of equal closeness applies. Sign convention: $V > 0$ favours the
    first-named model.
  - For M3 vs M1 (the exponential frontier versus the plain Poisson
    GLM) we report only the **AIC and BIC gaps**. The Poisson GLM is
    the $b \to \infty$ boundary of the exponential frontier, so the
    two families overlap at a boundary point; under such overlap
    Vuong's normal limit fails [@andrews2001], and the chi-squared LR
    is similarly invalid. The information-criterion gaps are
    boundary-agnostic and characterise the comparison cleanly.

```{r}
#| label: tbl-model-tests
#| tbl-cap: "Model-comparison statistics. V$_{32}$ is the Vuong (1989) $z$-statistic with two-sided $p$-value from the standard normal; LR$_2$, LR$_3$ (and LR$_4$ when M6 converges) are likelihood-ratio statistics with $\\chi^2_{df}$ reference distributions. For M3 vs M1 the boundary nesting precludes standard inference, so we report the AIC and BIC gaps (positive favouring M3) without a formal test. Critical values: standard normal $|z| > 1.96$ at the 5\\% level; $\\chi^2_1(0.95) = 3.84$; $\\chi^2_2(0.95) = 5.99$."

mc_df <- data.frame(
  Test         = c("M3 vs M1", "V$_{32}$: M3 vs M2",
                   "LR$_2$: M4 vs M3", "LR$_3$: M5 vs M4"),
  Type         = c("AIC/BIC gap (boundary)", "Vuong (non-nested)",
                   "Nested LR ($df=2$)", "Nested LR ($df=1$)"),
  `Comparison` = c("Poisson GLM vs Exp hom.", "PHN (M2) vs Exp hom.",
                   "Exp hom. vs Exp scaling", "Exp scaling vs +$\\log K^2$"),
  Statistic  = c(if (!is.null(ic31)) sprintf("$\\Delta$AIC=%.0f, $\\Delta$BIC=%.0f",
                                              ic31$aic_gap, ic31$bic_gap) else "---",
                  if (!is.null(v32) && !is.na(v32$V))   sprintf("%.2f", v32$V) else "---",
                  if (!is.na(lr2$stat)) sprintf("%.2f", max(lr2$stat, 0)) else "---",
                  if (!is.na(lr3$stat)) sprintf("%.2f", max(lr3$stat, 0)) else "---"),
  `p-value`  = c("---",
                  if (!is.null(v32) && !is.na(v32$pval)) sprintf("%.4f", v32$pval) else "---",
                  if (!is.na(lr2$pval)) sprintf("%.4f", lr2$pval) else "---",
                  if (!is.na(lr3$pval)) sprintf("%.4f", lr3$pval) else "---"),
  Decision   = c(if (!is.null(ic31) && ic31$aic_gap < 0) "Favours Exp hom." else "Favours Poisson",
                  if (!is.null(v32)) v32$decision else "N/A",
                  lr2$decision, lr3$decision),
  check.names = FALSE, stringsAsFactors = FALSE
)

if (m6_converged_sc && !is.null(lr4)) {
  mc_df <- rbind(mc_df, data.frame(
    Test         = "LR$_4$: M6 vs M3",
    Type         = "Nested LR ($df=1$)",
    Comparison   = "Exp hom. vs Gamma hom.",
    Statistic    = sprintf("%.2f", max(lr4$stat, 0, na.rm = TRUE)),
    `p-value`    = sprintf("%.4f", lr4$pval),
    Decision     = lr4$decision,
    check.names = FALSE, stringsAsFactors = FALSE
  ))
}

knitr::kable(
  mc_df, booktabs = TRUE, escape = FALSE,
  col.names = c("Test", "Type", "Comparison", "Statistic", "$p$-value", "Decision")
) |>
  kableExtra::kable_styling(latex_options = "scale_down")
```

The Gamma-shape diagnostic LR$_4$ in @tbl-model-tests bears directly on the
choice between exponential and Gamma inefficiency. With the scaling
structure in place, the free-$\alpha$ Gamma fit (M6) converges to
$\hat\alpha = 1.000$ at the exponential boundary; the test statistic is
LR$_4 = 0.00$ ($p = 1.00$), so the data provide no evidence against the
exponential restriction. This is consistent with the weak-identification
phenomenon documented in @sec-simulations: when the frontier mean
$\lambda_i$ is large relative to $b$ (as it is here, with mean patent
count $\approx 35$ and $\hat b \approx 0.52$), the curvature of the
log-likelihood in the $\alpha$ direction is shallow, and the sample carries
little information about the shape of the inefficiency density. The
exponential scaling frontier (M4/M5) is therefore retained as the preferred
specification.

The signs and magnitudes of the $\hat{\delta}$ estimates accord with prior
expectations. The capital coefficient
$\hat{\delta}_1 = `r round(delta1_hat, 3)`$
(SE $= `r round(delta1_se, 3)`$)
`r if (delta1_hat < 0) "is negative, consistent with the association between capital intensity and lower estimated inefficiency;" else "is positive, at odds with the prior expectation of an association between capital intensity and lower inefficiency;"`
Evaluated at the sample-mean inefficiency level, a one-unit increase in
$\log K_i$ shifts the predicted mean inefficiency by
$\hat{\delta}_1 \times \overline{\mathrm{E}(u_i)} = `r round(me_logk, 4)`$
log-units, a marginal effect derived from
$\partial\,\mathrm{E}(u_i)/\partial z_{i1} = \delta_1 \cdot \mathrm{E}(u_i)$
(see @eq-scaling-bi). The science-sector coefficient
$\hat{\delta}_2 = `r round(delta2_hat, 3)`$
(SE $= `r round(delta2_se, 3)`$) implies that firms in science-intensive
sectors show mean inefficiency `r round(100 * abs(1 - ineff_ratio), 1)`%
`r if (delta2_hat < 0) "lower" else "higher"` than comparable non-science
firms ($\exp(\hat{\delta}_2) = `r round(ineff_ratio, 3)`$), holding log
capital constant. This association is consistent with the superior absorptive
capacity and established IP workflows that characterise chemistry, electronics,
and scientific-instrument firms [@hallGrilichesHausman1986].

```{r}
#| label: fig-efficiency-scaling
#| fig-cap: "Distribution of firm-level efficiency scores under the preferred scaling model (solid) and the homogeneous exponential frontier (dashed). The scaling model accounts for capital stock and science-sector membership; the homogeneous model ignores these determinants. Vertical lines mark the respective sample means."

plot_dens <- data.frame(
  te    = c(sc$te_scaling, sc$te_homog),
  model = rep(c("Scaling (preferred)", "Homogeneous"), each = nrow(sc))
)
plot_dens <- plot_dens[is.finite(plot_dens$te), ]
means_dens <- aggregate(te ~ model, data = plot_dens, FUN = mean)

ggplot(plot_dens, aes(x = te, colour = model, linetype = model)) +
  geom_density(linewidth = 0.8, adjust = 1.2) +
  geom_vline(data = means_dens,
             aes(xintercept = te, colour = model, linetype = model),
             linewidth = 0.5, alpha = 0.7) +
  scale_colour_manual(
    values = c("Scaling (preferred)" = "steelblue", "Homogeneous" = "grey40"),
    name = NULL
  ) +
  scale_linetype_manual(
    values = c("Scaling (preferred)" = "solid", "Homogeneous" = "dashed"),
    name = NULL
  ) +
  labs(x = "Technical efficiency score", y = "Density") +
  theme_bw(base_size = 11) +
  theme(legend.position="bottom")
```

@fig-efficiency-scaling overlays the efficiency distributions from the
preferred scaling model and the homogeneous baseline. The scaling model
produces a distribution with
`r if (sd(sc$te_scaling, na.rm = TRUE) > sd(sc$te_homog, na.rm = TRUE)) "greater" else "lower"`
dispersion (SD $= `r round(sd(sc$te_scaling, na.rm = TRUE), 3)`$ vs
$`r round(sd(sc$te_homog, na.rm = TRUE), 3)`$ for the homogeneous model),
reflecting the additional cross-firm heterogeneity captured by the determinants.
The mean efficiency
`r if (te_mean_pref > te_mean_M3) "rises" else "falls"`
from `r te_mean_M3` (homogeneous) to `r te_mean_pref` (scaling), indicating
that ignoring the determinants
`r if (te_mean_pref > te_mean_M3) "under-estimates" else "over-estimates"`
average productive efficiency by conflating structural efficiency
advantages with random inefficiency.

```{r}
#| label: fig-eff-logk-scaling
#| fig-cap: "Firm efficiency scores against log capital stock, coloured by science-sector membership. LOESS smooths (span~=~0.75) are shown within each sector. Higher capital stock is negatively associated with estimated inefficiency, particularly in science-intensive industries."

plot_lk <- data.frame(
  te     = sc$te_scaling,
  logk   = sc$logk,
  sector = factor(ifelse(sc$sci == 1L, "Science", "Other"),
                  levels = c("Science", "Other"))
)
plot_lk <- plot_lk[is.finite(plot_lk$te), ]

ggplot(plot_lk, aes(x = logk, y = te, colour = sector)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_smooth(method = "loess", formula = y ~ x,
              span = 0.75, se = FALSE, linewidth = 1.0) +
  scale_colour_manual(
    values = c("Science" = "steelblue", "Other" = "grey40"), name = NULL
  ) +
  labs(x = expression(log~K[it]),
       y = expression(widehat(TE)[it])) +
  theme_bw(base_size = 11) +
  theme(legend.position="bottom")
```

```{r}
#| label: fig-eff-sector-scaling
#| fig-cap: "Efficiency score distributions by science-sector membership. Boxes span the interquartile range; the horizontal line is the median; whiskers extend to $1.5\\times$ IQR. Science-sector firms display higher median efficiency, consistent with the negative estimated $\\hat{\\delta}_2$."

plot_sec <- data.frame(
  te     = sc$te_scaling,
  sector = factor(ifelse(sc$sci == 1L, "Science", "Other"),
                  levels = c("Other", "Science"))
)
plot_sec <- plot_sec[is.finite(plot_sec$te), ]

ggplot(plot_sec, aes(x = sector, y = te)) +
  geom_jitter(width = 0.18, alpha = 0.18, size = 0.7, colour = "steelblue") +
  geom_boxplot(fill = NA, colour = "grey30",
               outlier.shape = NA, linewidth = 0.6, width = 0.40) +
  labs(x = NULL, y = expression(widehat(TE)[it])) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  theme_bw(base_size = 11)
```

@fig-eff-logk-scaling and @fig-eff-sector-scaling illuminate the two channels
through which the determinants shape the efficiency distribution. The positive
slope of the LOESS curves in @fig-eff-logk-scaling shows that capital
stock is negatively associated with estimated inefficiency (positively
associated with efficiency scores), consistent with $\hat{\delta}_1 < 0$.
Firms in science-intensive sectors (blue) show higher efficiency scores
than non-science firms (grey) at comparable capital levels, corroborating
the negative $\hat{\delta}_2$. @fig-eff-sector-scaling makes the
sector gap explicit: the median efficiency of science-sector firms exceeds that
of other firms by approximately
`r round(median(sc$te_scaling[sc$sci == 1L], na.rm = TRUE) - median(sc$te_scaling[sc$sci == 0L], na.rm = TRUE), 3)`
points, a practically meaningful difference relative to the sample mean of
`r te_mean_pref`.

Ignoring these determinants and fitting the homogeneous model (M3) absorbs the
sector- and capital-induced heterogeneity into the single rate parameter $b$,
compressing the efficiency distribution. The firms for which the gap between
scaling and homogeneous scores is largest are precisely the capital-intensive
science-sector observations that the determinants identify as nearest to the
frontier: the homogeneous model systematically under-rates their efficiency,
masking the complementarity between physical capital and innovation capacity.
This pattern mirrors the simulation evidence of @sec-sim-scaling, where
ignoring strong inefficiency determinants was shown to bias the estimated
overall inefficiency level upward. The results support the recommendation of
[@wangSchmidt2002] that one-step estimation with firm-level covariates is
preferable to the homogeneous frontier whenever economic theory motivates a
heterogeneous inefficiency distribution.
