# Empirical Application {#sec-empirical}

```{r}
#| label: setup-06
#| include: false
#| cache: false
library(countSFA)
library(ggplot2)
library(kableExtra)
library(data.table)
options(knitr.table.format = if (knitr::is_latex_output()) "latex" else "html")
```

```{r}
#| label: cwa-run
#| cache: true
#| include: false
#| error: true
# The CWA application script caches each model fit and the score posteriors
# individually under results/, so re-runs are incremental. The consolidated
# results/cwa_fits.rds object is written at the end of the script.
if (!file.exists("results/cwa_fits.rds")) {
  source("code/empirical/cwa_application.R")
}
```

```{r}
#| label: cwa-load
#| include: false
emp <- readRDS("results/cwa_fits.rds")

dat        <- emp$data
y          <- emp$y
X          <- emp$X
Z          <- emp$Z
fit_pois   <- emp$fit_pois
fit_exp    <- emp$fit_exp
fit_gam    <- emp$fit_gam
fit_M4     <- emp$fit_M4
fit_M5     <- emp$fit_M5
scores_exp <- emp$scores_exp
scores_gam <- emp$scores_gam
scores_M4  <- emp$scores_M4
scores_M5  <- emp$scores_M5
lr_tests   <- emp$lr_tests
n_obs      <- length(y)
n_X        <- ncol(X)
n_Z        <- ncol(Z)

# Convergence and preferred-model selection
exp_ok <- !is.null(fit_exp) && isTRUE(fit_exp$convergence == 0L) &&
          is.finite(fit_exp$loglik) && fit_exp$loglik > -1e10
gam_ok <- !is.null(fit_gam) && isTRUE(fit_gam$convergence == 0L) &&
          is.finite(fit_gam$loglik) && fit_gam$loglik > -1e10
M4_ok  <- !is.null(fit_M4)  && isTRUE(fit_M4$convergence  == 0L) &&
          is.finite(fit_M4$loglik)  && fit_M4$loglik  > -1e10
M5_ok  <- !is.null(fit_M5)  && isTRUE(fit_M5$convergence  == 0L) &&
          is.finite(fit_M5$loglik)  && fit_M5$loglik  > -1e10

# Prefer the most general convergent specification for the posterior scores:
# M5 (Gamma + scaling) > M4 (Exp + scaling) > M3 (Gamma homog.) > M2 (Exp homog.)
preferred_fit <- if (M5_ok) fit_M5 else
                 if (M4_ok) fit_M4 else
                 if (gam_ok) fit_gam else fit_exp
preferred_label <- if (M5_ok) "Gamma scaling" else
                   if (M4_ok) "exponential scaling" else
                   if (gam_ok) "Gamma" else "exponential"

# Detection / compliance posteriors. Posterior of exp(-u_i) under the
# cost-frontier orientation: this is the ratio of the latent feasible
# minimum violation count to the observed conditional mean — interpretable
# as a compliance-efficiency score in [0, 1]. The scores must match the
# preferred specification: the scaling models (M4/M5) carry the heterogeneous
# rate b_i = b * exp(-z_i' delta), so their posteriors differ from the
# homogeneous M2/M3 scores.
preferred_scores <- switch(
  preferred_label,
  "Gamma scaling"       = scores_M5,
  "exponential scaling" = scores_M4,
  "Gamma"               = scores_gam,
  scores_exp)
# Fall back to the homogeneous posteriors if the scaling scores are unavailable.
if (is.null(preferred_scores))
  preferred_scores <- if (gam_ok) scores_gam else scores_exp

# Inline number helpers
ll <- function(f) {
  if (is.null(f)) return(NA_real_)
  if (!is.null(f$loglik)) return(f$loglik)
  if (inherits(f, "glm"))  return(as.numeric(logLik(f)))
  NA_real_
}
ll_M1 <- ll(fit_pois); ll_M2 <- ll(fit_exp); ll_M3 <- ll(fit_gam)
ll_M4 <- ll(fit_M4);   ll_M5 <- ll(fit_M5)
```

## Data and Cost-Frontier Framing {#sec-emp-data}

We apply the Poisson stochastic cost frontier to U.S. Environmental Protection Agency (EPA) National Pollutant Discharge Elimination System (NPDES) facility data. The outcome is constructed from $n = `r format(nrow(dat), big.mark = ",")`$ major-permit facility observations, where for each facility $i$ the count

$$
Y_i \;=\; \text{number of quarters in non-compliance over the 13-quarter rolling window}
$$

is constructed from the EPA ECHO Quarterly Non-Compliance Report. A "quarter in non-compliance" is any quarter in which the facility was in reportable non-compliance (RNC), in violation, or in significant non-compliance (SNC) status. By construction $Y_i \in \{0, 1, \ldots, 13\}$. We frame non-compliance as an *undesirable* output that the regulated facility seeks to minimise: every quarter in non-compliance triggers expected enforcement actions, civil penalties, and reputational costs, so the count behaves economically like the dollar value of compliance failures rather than a desirable output like patents or units produced. Accordingly we adopt the cost-frontier orientation of @sec-model throughout. Writing $a_i = \mathbf{x}_i'\boldsymbol{\beta}$ for the log of the *feasible minimum* expected non-compliance count given facility characteristics,

$$
Y_i \mid u_i,\,\mathbf{x}_i \;\sim\; \mathrm{Poisson}\!\left(\exp(a_i + u_i)\right),
\qquad u_i \;\sim\; \mathrm{Gamma}(\alpha,\, b),\quad u_i \geq 0,
$$ {#eq-cwa-dgp}

so that observed violations are inflated above the frontier by the multiplicative factor $\exp(u_i)\geq 1$. The frontier $\lambda_i = e^{a_i}$ is the conditional expected count a perfectly-managed facility (one with $u_i = 0$) would attain given the size, sector, location, and regulatory context $\mathbf{x}_i$. Departures above this floor are attributed to managerial inefficiency in compliance — gaps in monitoring, slow response to deviations, under-investment in compliance staff, or weak internal reporting — captured by $u_i$. The conditional efficiency score $\mathrm{E}[e^{-u_i}\mid y_i, \mathbf{x}_i] \in (0,1]$ measures the proportion of the irreducible benchmark that facility $i$ achieves: $1$ if the facility operates at the feasible minimum, lower otherwise.

```{r}
#| label: tbl-cwa-summary
#| tbl-cap: !expr paste0("Summary statistics for the EPA NPDES major-permit panel (n = ", format(n_obs, big.mark = ","), " facilities). The outcome $Y_i$ counts quarters in non-compliance over a rolling 13-quarter window. Substantial overdispersion (variance/mean $\\approx ", round(var(y)/mean(y), 1), "$) and a large zero share are consistent with a cost-frontier specification in which compliance is intermittently failed by a subset of facilities while most operate at or near the feasible minimum.")

q <- quantile(y, c(.5, .75, .9, .95, .99))
sum_tbl <- data.frame(
  Statistic = c("n", "Mean", "Variance", "Variance/Mean",
                "Median", "75th pct.", "90th pct.", "95th pct.", "99th pct.",
                "Maximum", "Zero share"),
  Value = c(format(n_obs, big.mark = ","),
            sprintf("%.3f", mean(y)),
            sprintf("%.3f", var(y)),
            sprintf("%.2f", var(y)/mean(y)),
            sprintf("%.0f", q[1]),
            sprintf("%.0f", q[2]),
            sprintf("%.0f", q[3]),
            sprintf("%.0f", q[4]),
            sprintf("%.0f", q[5]),
            sprintf("%d", max(y)),
            sprintf("%.1f%%", 100 * mean(y == 0))),
  stringsAsFactors = FALSE
)

kable(sum_tbl, booktabs = TRUE, escape = FALSE,
      col.names = c("Statistic", "Value"),
      align = c("l", "r")) |>
  kable_styling(latex_options = "scale_down")
```

@tbl-cwa-summary documents the salient features of the count distribution. The sample mean is `r round(mean(y), 2)` non-compliant quarters per facility, but the variance of `r round(var(y), 2)` yields a variance-to-mean ratio of `r round(var(y)/mean(y), 1)`, far above the equidispersed Poisson benchmark of 1. The zero share is `r round(100*mean(y == 0), 1)`%, dominating the lower end of the distribution, while the upper tail is heavy: `r round(100 * mean(y >= 6), 1)`% of facilities accumulate six or more non-compliant quarters. The cost-frontier specification of @eq-cwa-dgp accommodates both features simultaneously: facilities at or near the feasible minimum produce mostly zeros under the Poisson conditional, while inefficient facilities ($u_i$ large) generate the heavy upper tail through the multiplicative inflation $\exp(u_i)$.

## Empirical Specification {#sec-emp-spec}

The frontier mean is parameterised log-linearly, $a_i = \mathbf{x}_i'\boldsymbol{\beta}$, with `r n_X` covariates in $\mathbf{x}_i$ chosen to capture facility characteristics that shift the *irreducible* expected non-compliance count under best practice. These include the 2-digit NAICS sector (water and wastewater treatment, NAICS 22, is the baseline; chemicals, mining, food manufacturing, and other heavy-discharge sectors are entered as dummies); ten EPA regional dummies to absorb cross-state regulatory baselines and ambient water-quality conditions; the log of one plus total inspections, a size-and-risk proxy that loads positively on the latent rate because larger or higher-risk facilities receive more inspections and have more opportunities to be cited; the percent minority and log population density of the surrounding census block group; and indicator flags for Indian Country location, impaired-water-receiving status, and co-regulation under the Clean Air Act, RCRA, the Toxics Release Inventory, and the Greenhouse Gas Reporting Program. The co-regulation flags proxy multimedia regulatory complexity: facilities that face simultaneous oversight in multiple media tend to operate larger, more chemically-diverse processes with higher *intrinsic* non-compliance potential.

The scaling determinants $\mathbf{z}_i$ capture facility-level drivers of inefficiency in compliance management. We model the inefficiency rate as $b_i = b\,\exp(-\mathbf{z}_i'\boldsymbol{\delta})$, so that $\mathrm{E}[u_i\mid\mathbf{z}_i] = \alpha\exp(\mathbf{z}_i'\boldsymbol{\delta})/b$ and a positive component $\delta_j$ raises facility $i$'s expected inefficiency. Four determinants are entered:

- *Log days since last inspection* ($z_1$): a measure of regulatory attention lapse. Facilities that have not been inspected recently lack the discipline of imminent scrutiny; the cost-orientation prediction is $\delta_1 > 0$.
- *Log formal enforcement actions* ($z_2$): a stock measure of past regulatory action against the facility. The interpretation is two-sided. A strong deterrence channel would predict $\delta_2 < 0$ (past enforcement teaches the facility to comply going forward); a selection channel would predict $\delta_2 > 0$ (formal enforcement is targeted at chronic violators, so the stock is a marker for inefficient compliance management).
- *Log total penalty dollars* ($z_3$): a stock measure of monetary punishment. Same two-sided logic as $z_2$.
- *Percent minority in surrounding block group* ($z_4$): an environmental-justice proxy. The empirical literature documents that facilities in high-minority communities face systematically less monitoring scrutiny relative to comparable facilities in low-minority communities [@konisky2009; @grayShadbegian2004], which would map to higher operational latitude for compliance shortcuts and a positive $\delta_4$. The percent-minority variable also enters $\mathbf{x}_i$ at the frontier; identification of $(\beta_{\text{minority}},\, \delta_{\text{minority}})$ comes from the multiplicative structure of the cost-frontier Gamma conditional, not from an exclusion restriction, and is discussed in @sec-emp-determinants.

```{r}
#| label: tbl-cwa-Z-summary
#| tbl-cap: "Summary statistics for the four scaling-model determinants entering $\\mathbf{z}_i$. The log transformations stabilise the right tails of the inspection-gap and enforcement-stock variables."
Z_summary <- data.frame(
  Variable = c("log(1 + days since last inspection / 365)",
               "log(1 + formal enforcement actions)",
               "log(1 + total penalties \\$)",
               "Percent minority (0--100)"),
  Mean = sapply(seq_len(ncol(Z)), function(j) sprintf("%.2f", mean(Z[, j]))),
  SD   = sapply(seq_len(ncol(Z)), function(j) sprintf("%.2f", sd(Z[, j]))),
  Min  = sapply(seq_len(ncol(Z)), function(j) sprintf("%.2f", min(Z[, j]))),
  Max  = sapply(seq_len(ncol(Z)), function(j) sprintf("%.2f", max(Z[, j]))),
  stringsAsFactors = FALSE
)
kable(Z_summary, booktabs = TRUE, escape = FALSE,
      col.names = c("Determinant ($z_j$)", "Mean", "SD", "Min", "Max"),
      align = c("l", "r", "r", "r", "r")) |>
  kable_styling(latex_options = "scale_down")
```

## Estimation Results {#sec-emp-results}

Five specifications are estimated. **M1** is a standard Poisson GLM with no frontier structure: $Y_i\mid\mathbf{x}_i \sim \mathrm{Poisson}(\exp(\mathbf{x}_i'\boldsymbol{\beta}))$. It is the misspecified benchmark, since it ignores the one-sided inefficiency. **M2** is the cost-frontier model with exponential inefficiency ($\alpha = 1$ fixed). **M3** is the cost-frontier model with free $\alpha$, the full Gamma. **M4** adds the scaling block to M2: $u_i\sim\mathrm{Exp}(b\exp(-\mathbf{z}_i'\boldsymbol{\delta}))$. **M5** is the most general specification, the Gamma cost frontier with scaling.

```{r}
#| label: tbl-cwa-main
#| tbl-cap: !expr paste0("Maximum-likelihood estimates of the cost-frontier specifications, $n = ", format(n_obs, big.mark = ","), "$ facilities. SEs in parentheses. M1: Poisson GLM; M2: exponential cost frontier; M3: Gamma cost frontier; M4: exponential + scaling on $\\mathbf{z}_i$; M5: Gamma + scaling. The frontier coefficient vector $\\boldsymbol{\\beta}$ contains $", n_X, "$ entries; the table reports a focused subset of substantively important coefficients. Standard errors for $\\hat{b}$ and $\\hat{\\alpha}$ use the delta method on the corresponding $\\log$-parameters; standard errors for $\\hat{\\delta}$ are read directly from the inverse Hessian.")

# Helper: format value with SE
fmt_v <- function(v, s) {
  if (is.null(v) || is.na(v)) return("---")
  if (is.null(s) || is.na(s)) return(sprintf("%.3f", v))
  sprintf("%.3f (%.3f)", v, s)
}

# Indexing the small set of named X covariates we report (positions in X)
key_cols <- c("(Intercept)", "log_insp", "FAC_PERCENT_MINORITY",
              "log_pop_den", "AIR_FLAG_Y", "RCRA_FLAG_Y", "TRI_FLAG_Y",
              "GHG_FLAG_Y")
key_cols <- intersect(key_cols, colnames(X))
key_lab  <- c("(Intercept)"           = "Intercept",
              "log_insp"              = "$\\log(1+\\text{inspections})$",
              "FAC_PERCENT_MINORITY"  = "Percent minority",
              "log_pop_den"           = "$\\log$ pop. density",
              "AIR_FLAG_Y"            = "CAA flag",
              "RCRA_FLAG_Y"           = "RCRA flag",
              "TRI_FLAG_Y"            = "TRI flag",
              "GHG_FLAG_Y"            = "GHG flag")

# Extract (beta, SE) from each fit at the named columns
extract_pois <- function(fit) {
  if (is.null(fit)) return(setNames(rep("---", length(key_cols)), key_cols))
  cf <- coef(fit)
  se <- sqrt(diag(vcov(fit)))
  # GLM column names are prefixed with "X"
  cf_names <- sub("^X", "", names(cf))
  out <- setNames(rep("---", length(key_cols)), key_cols)
  for (cc in key_cols) {
    j <- match(cc, cf_names)
    if (!is.na(j)) out[cc] <- fmt_v(cf[j], se[j])
  }
  out
}
extract_frontier <- function(fit) {
  if (is.null(fit)) return(setNames(rep("---", length(key_cols)), key_cols))
  cf <- fit$coefficients
  se <- fit$se
  out <- setNames(rep("---", length(key_cols)), key_cols)
  for (cc in key_cols) {
    j <- match(cc, colnames(X))
    if (!is.na(j)) out[cc] <- fmt_v(cf[j], se[j])
  }
  out
}
extract_scaling <- function(fit) {
  if (is.null(fit)) return(setNames(rep("---", length(key_cols)), key_cols))
  cf <- fit$beta
  vv <- if (!is.null(fit$vcov)) sqrt(pmax(diag(fit$vcov), 0)) else
        rep(NA_real_, length(cf))
  out <- setNames(rep("---", length(key_cols)), key_cols)
  for (cc in key_cols) {
    j <- match(cc, colnames(X))
    if (!is.na(j)) out[cc] <- fmt_v(cf[j], vv[j])
  }
  out
}

m1_row <- extract_pois(fit_pois)
m2_row <- extract_frontier(fit_exp)
m3_row <- extract_frontier(fit_gam)
m4_row <- extract_scaling(fit_M4)
m5_row <- extract_scaling(fit_M5)

# b, alpha, delta rows
b_row    <- c(M1 = "---",
              M2 = fmt_v(fit_exp$b, fit_exp$se_b),
              M3 = fmt_v(fit_gam$b, fit_gam$se_b),
              M4 = if (!is.null(fit_M4)) fmt_v(fit_M4$b, sqrt(pmax(diag(fit_M4$vcov), 0))[n_X + 1L] * fit_M4$b) else "---",
              M5 = if (!is.null(fit_M5)) fmt_v(fit_M5$b, sqrt(pmax(diag(fit_M5$vcov), 0))[n_X + 1L] * fit_M5$b) else "---")
alpha_row <- c(M1 = "---", M2 = "1 (fixed)",
               M3 = fmt_v(fit_gam$alpha, fit_gam$se_alpha),
               M4 = "1 (fixed)",
               M5 = if (!is.null(fit_M5)) {
                 sqrt_diag <- sqrt(pmax(diag(fit_M5$vcov), 0))
                 fmt_v(fit_M5$alpha, sqrt_diag[n_X + 2L] * fit_M5$alpha)
               } else "---")

# Scaling delta block (M4 and M5)
delta_fmt <- function(fit, j) {
  if (is.null(fit) || is.null(fit$delta)) return("---")
  vv <- sqrt(pmax(diag(fit$vcov), 0))
  # index of delta_j depends on whether alpha is free
  fixed_alpha <- (fit$dist == "exponential")
  idx <- if (fixed_alpha) n_X + 1L + j else n_X + 2L + j
  fmt_v(fit$delta[j], vv[idx])
}
delta_rows <- lapply(seq_len(n_Z), function(j) {
  c(M1 = "---", M2 = "---", M3 = "---",
    M4 = delta_fmt(fit_M4, j), M5 = delta_fmt(fit_M5, j))
})

# Build full coefficient table
build_col <- function(col_extractor) col_extractor
all_x_rows <- mapply(function(cc) {
  c(M1 = m1_row[cc], M2 = m2_row[cc], M3 = m3_row[cc],
    M4 = m4_row[cc], M5 = m5_row[cc])
}, key_cols, SIMPLIFY = FALSE)

# Compose table
panel_X <- do.call(rbind, lapply(seq_along(key_cols), function(i) {
  c(key_lab[key_cols[i]], all_x_rows[[i]])
}))
panel_b <- c("$\\hat{b}$", b_row)
panel_a <- c("$\\hat{\\alpha}$", alpha_row)
panel_delta <- do.call(rbind, lapply(seq_len(n_Z), function(j) {
  c(paste0("$\\hat{\\delta}_", j, "$ (", colnames(Z)[j], ")"), delta_rows[[j]])
}))
panel_fit <- rbind(
  c("Log-likelihood",
    sprintf("%.0f", ll_M1), sprintf("%.0f", ll_M2),
    if (is.finite(ll_M3)) sprintf("%.0f", ll_M3) else "---",
    if (is.finite(ll_M4)) sprintf("%.0f", ll_M4) else "---",
    if (is.finite(ll_M5)) sprintf("%.0f", ll_M5) else "---"),
  c("AIC",
    sprintf("%.0f", AIC(fit_pois)),
    sprintf("%.0f", fit_exp$AIC),
    if (!is.null(fit_gam)) sprintf("%.0f", fit_gam$AIC) else "---",
    if (!is.null(fit_M4))  sprintf("%.0f", fit_M4$AIC)  else "---",
    if (!is.null(fit_M5))  sprintf("%.0f", fit_M5$AIC)  else "---")
)

tbl_main <- rbind(panel_X, panel_b, panel_a, panel_delta, panel_fit)
colnames(tbl_main) <- c(" ", "M1: Poisson", "M2: Exp", "M3: Gamma",
                        "M4: Exp+Z", "M5: Gamma+Z")

# Render with section dividers (pack_rows)
n_X_panel <- nrow(panel_X)
n_dist    <- 2L                  # b, alpha rows
n_d       <- nrow(panel_delta)
end_X     <- n_X_panel
end_dist  <- end_X + n_dist
end_d     <- end_dist + n_d

kable(tbl_main, booktabs = TRUE, escape = FALSE, row.names = FALSE,
      align = c("l", rep("r", 5))) |>
  pack_rows("Frontier $\\\\boldsymbol{\\\\beta}$ (selected)",
            1, end_X, escape = FALSE) |>
  pack_rows("Inefficiency distribution",
            end_X + 1L, end_dist, escape = FALSE) |>
  pack_rows("Inefficiency determinants $\\\\boldsymbol{\\\\delta}$",
            end_dist + 1L, end_d, escape = FALSE) |>
  pack_rows("Fit",
            end_d + 1L, nrow(tbl_main), escape = FALSE) |>
  kable_styling(latex_options = "scale_down", font_size = 8) |>
  footnote(general = "Sector and EPA-region fixed effects included but not reported.",
           general_title = "", footnote_as_chunk = TRUE, escape = FALSE)
```

@tbl-cwa-main reports the key estimates. The Poisson GLM (M1) has the lowest log-likelihood by a wide margin; moving to the exponential cost frontier (M2) raises the log-likelihood by `r format(round(ll_M2 - ll_M1), big.mark = ",")` log-units at the cost of one additional parameter (the rate $b$). The Gamma extension (M3) further raises the log-likelihood by `r if (is.finite(ll_M3)) format(round(ll_M3 - ll_M2), big.mark = ",") else "---"` units, indicating that allowing an interior mode in the inefficiency density adds meaningful structure relative to the exponential restriction. Adding the scaling block in M4 and M5 raises the log-likelihood by `r if (is.finite(ll_M4) && is.finite(ll_M2)) format(round(ll_M4 - ll_M2), big.mark = ",") else "---"` and `r if (is.finite(ll_M5) && is.finite(ll_M3)) format(round(ll_M5 - ll_M3), big.mark = ",") else "---"` units respectively against the corresponding homogeneous nest, confirming that the four inefficiency determinants in $\mathbf{z}_i$ contain substantial information about compliance heterogeneity beyond what the frontier covariates capture.

```{r}
#| label: tbl-cwa-lr
#| tbl-cap: "Nested likelihood-ratio tests across the five specifications. Critical values: $\\chi^2_1(0.95) = 3.84$, $\\chi^2_4(0.95) = 9.49$. All four tests reject the restricted model at the 1\\% level (when the unrestricted alternative converged)."

lr_df <- data.frame(
  Test = c("M3 vs M2", "M4 vs M2", "M5 vs M3", "M5 vs M4"),
  Restriction = c("$\\alpha = 1$",
                  "$\\boldsymbol{\\delta} = \\mathbf{0}$",
                  "$\\boldsymbol{\\delta} = \\mathbf{0}$",
                  "$\\alpha = 1$"),
  `df` = c("1", as.character(n_Z), as.character(n_Z), "1"),
  Statistic = sapply(c("M3_vs_M2", "M4_vs_M2", "M5_vs_M3", "M5_vs_M4"),
    function(k) {
      t <- lr_tests[[k]]
      if (is.null(t) || is.na(t$stat)) "---"
      else sprintf("%.1f", max(t$stat, 0))
    }),
  `p-value` = sapply(c("M3_vs_M2", "M4_vs_M2", "M5_vs_M3", "M5_vs_M4"),
    function(k) {
      t <- lr_tests[[k]]
      if (is.null(t) || is.na(t$pval)) "---"
      else if (t$pval < 1e-4) "$<10^{-4}$"
      else sprintf("%.4f", t$pval)
    }),
  check.names = FALSE, stringsAsFactors = FALSE
)
kable(lr_df, booktabs = TRUE, escape = FALSE,
      align = c("l", "l", "r", "r", "r")) |>
  kable_styling(latex_options = "scale_down")
```

The likelihood-ratio tests in @tbl-cwa-lr formalise the model-comparison story. The Gamma-shape extension is rejected by the data against the exponential restriction (M3 vs M2): the data prefer an interior-mode inefficiency density. The scaling block is rejected against both the exponential and the Gamma homogeneous baseline (M4 vs M2, M5 vs M3) — the determinants explain variation in inefficiency that the homogeneous parameter $b$ cannot accommodate. The Gamma-vs-exponential comparison conditional on scaling (M5 vs M4) similarly rejects, supporting M5 as the preferred specification.

::: {.remark title="Identification under dual-use covariates"}
Percent minority enters both the frontier $\mathbf{x}_i$ and the scaling block $\mathbf{z}_i$. There is no exclusion restriction. Identification comes from the multiplicative structure of the cost-frontier Gamma conditional: $\beta_{\text{minority}}$ shifts the *level* of the latent frontier $\lambda_i$, whereas $\delta_{\text{minority}}$ shifts the *rate* parameter $b_i$ of the inefficiency distribution and hence the dispersion as well as the location of $u_i\mid\mathbf{x}_i$. The two parameters affect different conditional moments and are therefore separately identified at any sample size at which the higher moments of the count are informative. The Monte Carlo experiments of @sec-sim-scaling confirm that the one-step MLE recovers $(\boldsymbol{\beta},\boldsymbol{\delta})$ with near-nominal coverage when an overlapping covariate appears in both blocks under the cost-frontier DGP.
:::

## Compliance Efficiency Posteriors {#sec-emp-efficiency}

The posterior expectation $\widehat{\mathrm{TE}}_i = \mathrm{E}[e^{-u_i}\mid y_i, \mathbf{x}_i]$ is the conditional probability that facility $i$ operates at the feasible compliance minimum, given its observed history $y_i$ and characteristics $\mathbf{x}_i$. Under the cost-orientation specification @eq-cwa-dgp this score is well-defined for every $b > 0$ (@sec-app-condmoments), even on parameter regions where the marginal mean of $Y_i$ would diverge. @fig-cwa-eff plots the histogram of $\widehat{\mathrm{TE}}_i$ across all `r format(n_obs, big.mark = ",")` facilities under the preferred `r preferred_label` specification.

```{r}
#| label: fig-cwa-eff
#| fig-cap: !expr paste0("Distribution of compliance-efficiency scores $\\widehat{\\mathrm{TE}}_i = \\mathrm{E}[e^{-u_i}\\mid y_i,\\mathbf{x}_i]$ across the ", format(n_obs, big.mark = ","), " NPDES major-permit facilities, computed under the preferred ", preferred_label, " cost-frontier specification. Scores of 1 indicate operation at the irreducible compliance minimum; lower scores indicate progressively larger management inefficiency. The dashed line marks the sample mean.")
#| fig-width: 6
#| fig-height: 3.6

te <- preferred_scores$eff_score
te <- te[is.finite(te)]

ggplot(data.frame(te = te), aes(x = te)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = mean(te), linetype = "dashed",
             colour = "firebrick", linewidth = 0.7) +
  annotate("text", x = mean(te), y = Inf,
           label = paste0("Mean = ", sprintf("%.3f", mean(te))),
           hjust = -0.05, vjust = 1.6, size = 3.2, colour = "firebrick") +
  labs(x = expression(widehat(TE)[i] == E*"["*e^{-u[i]}*"|"*y[i]*","*bold(x)[i]*"]"),
       y = "Number of facilities") +
  theme_bw(base_size = 11)
```

The mean compliance-efficiency score is `r sprintf("%.3f", mean(preferred_scores$eff_score, na.rm = TRUE))`, indicating that the average facility attains only about `r sprintf("%.0f%%", 100 * mean(preferred_scores$eff_score, na.rm = TRUE))` of the irreducible compliance benchmark implied by its size, sector, and regulatory exposure — equivalently, observed non-compliance runs well above the feasible minimum for most units. The distribution is concentrated at low scores with a thin right tail: the bulk of facilities cluster near the lower end, while a minority reach modestly higher efficiency, reflecting persistent inefficiency in compliance management beyond what facility characteristics alone would imply.

```{r}
#| label: fig-cwa-eff-by-mins
#| fig-cap: "Compliance-efficiency score against percent minority in the surrounding census block group. Each point is one facility; the LOESS smooth (span 0.6) summarises the conditional mean. A downward slope is consistent with $\\hat{\\delta}_{\\text{minority}} > 0$ in @tbl-cwa-main: facilities in higher-minority communities operate further above the feasible compliance minimum, after controlling for sector, size, regulatory exposure, and inspection history."
#| fig-width: 6
#| fig-height: 3.6

plot_df <- data.frame(te = preferred_scores$eff_score,
                      pct = dat$FAC_PERCENT_MINORITY)
plot_df <- plot_df[is.finite(plot_df$te) & is.finite(plot_df$pct), ]

ggplot(plot_df, aes(x = pct, y = te)) +
  geom_point(alpha = 0.10, size = 0.6, colour = "steelblue") +
  geom_smooth(method = "loess", span = 0.6, se = FALSE,
              colour = "firebrick", linewidth = 0.9, formula = y ~ x) +
  labs(x = "Percent minority (block group)",
       y = expression(widehat(TE)[i])) +
  theme_bw(base_size = 11)
```

@fig-cwa-eff-by-mins displays the relationship between the compliance-efficiency score and the percent minority in the surrounding census block group. The negative LOESS slope, in combination with the positive scaling coefficient $\hat{\delta}_{\text{minority}}$ in @tbl-cwa-main, supports the interpretation that high-minority communities host facilities with systematically larger compliance inefficiency, even after the frontier specification absorbs any direct effect of community demographics on the latent baseline rate.

## Inefficiency Determinants {#sec-emp-determinants}

The scaling coefficients $\hat{\boldsymbol{\delta}}$ in @tbl-cwa-main carry direct economic content under the cost-frontier orientation: a positive $\hat{\delta}_j$ raises facility-level inefficiency, $\mathrm{E}[u_i\mid\mathbf{z}_i] = \alpha\exp(\mathbf{z}_i'\boldsymbol{\delta})/b$, by an amount proportional to the determinant. The marginal effect of $z_{ij}$ on $\mathrm{E}[u_i]$ is

$$
\frac{\partial \mathrm{E}[u_i \mid \mathbf{z}_i]}{\partial z_{ij}}
\;=\; \delta_j\,\mathrm{E}[u_i \mid \mathbf{z}_i],
$$ {#eq-cwa-marginal}

a *proportional* shift in expected inefficiency. The signs in @tbl-cwa-main reproduce the priors articulated in @sec-emp-spec.

```{r}
#| label: tbl-cwa-delta-marginals
#| tbl-cap: "Inefficiency-determinant estimates and their marginal effects on $\\mathrm{E}[u_i]$ evaluated at the sample-mean inefficiency level under the preferred specification. The marginal effect $\\delta_j \\times \\overline{\\mathrm{E}[u_i]}$ translates the proportional effect into expected-non-compliance-quarter units."

# Sample-mean E[u_i] under preferred fit
mean_Eu <- if (preferred_label == "Gamma scaling") {
  bi <- preferred_fit$b * exp(-drop(Z %*% preferred_fit$delta))
  mean(preferred_fit$alpha / bi)
} else if (preferred_label == "exponential scaling") {
  bi <- preferred_fit$b * exp(-drop(Z %*% preferred_fit$delta))
  mean(1 / bi)
} else if (preferred_label == "Gamma") {
  preferred_fit$alpha / preferred_fit$b
} else {
  1 / preferred_fit$b
}

delta_hat <- if (!is.null(preferred_fit$delta)) preferred_fit$delta else
             rep(NA_real_, n_Z)
se_delta_hat <- if (!is.null(preferred_fit$vcov)) {
  vv <- sqrt(pmax(diag(preferred_fit$vcov), 0))
  fixed_alpha <- preferred_fit$dist == "exponential"
  idx <- if (fixed_alpha) (n_X + 2L):(n_X + 1L + n_Z) else
                          (n_X + 3L):(n_X + 2L + n_Z)
  if (length(idx) == n_Z) vv[idx] else rep(NA_real_, n_Z)
} else rep(NA_real_, n_Z)

delta_tbl <- data.frame(
  Determinant = c("$\\log(1+\\text{days since last insp.}/365)$",
                  "$\\log(1+\\text{formal enforcement actions})$",
                  "$\\log(1+\\text{total penalty \\$})$",
                  "Percent minority"),
  `$\\hat{\\delta}_j$` = sapply(seq_len(n_Z),
                              function(j) sprintf("%.3f", delta_hat[j])),
  `SE` = sapply(seq_len(n_Z),
                function(j) sprintf("%.3f", se_delta_hat[j])),
  `$\\partial\\mathrm{E}[u_i]/\\partial z_{ij}$` =
    sapply(seq_len(n_Z),
           function(j) sprintf("%+.4f", delta_hat[j] * mean_Eu)),
  check.names = FALSE, stringsAsFactors = FALSE
)
kable(delta_tbl, booktabs = TRUE, escape = FALSE,
      align = c("l", "r", "r", "r")) |>
  kable_styling(latex_options = "scale_down") |>
  footnote(general = paste0("Marginal effects evaluated at the sample mean $\\\\overline{\\\\mathrm{E}[u_i]} = ",
                            sprintf("%.3f", mean_Eu), "$ under the preferred ",
                            preferred_label, " specification."),
           general_title = "", footnote_as_chunk = TRUE, escape = FALSE)
```

Contrary to the attention-lapse prior of @sec-emp-spec, the inspection-gap effect $\hat{\delta}_1 = `r sprintf("%.3f", delta_hat[1])`$ (SE $`r sprintf("%.3f", se_delta_hat[1])`$) is small, negative, and not significant at the 5% level: a longer interval since the last inspection is not associated with greater compliance inefficiency in this panel. A plausible reading is endogenous inspection timing — regulators direct scrutiny toward facilities they already suspect of weak compliance, so recently-inspected units are selected toward higher inefficiency and the raw inspection gap absorbs this reverse selection rather than an attention-lapse effect, the same selection logic that governs the enforcement-stock determinants below. The enforcement-stock variables yield estimates consistent with the *selection* interpretation rather than the deterrence interpretation: both $\hat{\delta}_2$ (formal actions) and $\hat{\delta}_3$ (total penalties) are positive, indicating that the historical stock of enforcement is a marker of chronic non-compliance rather than a deterrent against future violations — at least at the average regulator–facility margin observed in this panel. The percent-minority coefficient $\hat{\delta}_4 = `r sprintf("%.3f", delta_hat[4])`$ (SE $`r sprintf("%.3f", se_delta_hat[4])`$) is positive and statistically distinct from zero, consistent with the environmental-justice literature on systematic under-monitoring of high-minority communities translating into greater latitude for compliance shortcuts at facilities in those communities, after the frontier has absorbed the direct effect of demographics on baseline violation propensity.

```{r}
#| label: fig-cwa-eff-by-insp
#| fig-cap: "Compliance-efficiency score against the log inspection-gap variable $z_1 = \\log(1 + \\text{days since last inspection}/365)$. Each point is a facility; the LOESS smooth captures the conditional mean. The faint upward slope mirrors the small, negative $\\hat{\\delta}_1$ in @tbl-cwa-delta-marginals: in this panel, facilities further from their last inspection are not estimated to drift further from the compliance frontier."
#| fig-width: 6
#| fig-height: 3.6

plot_df2 <- data.frame(te = preferred_scores$eff_score,
                       z1 = Z[, "log_days_insp"])
plot_df2 <- plot_df2[is.finite(plot_df2$te) & is.finite(plot_df2$z1), ]

ggplot(plot_df2, aes(x = z1, y = te)) +
  geom_point(alpha = 0.10, size = 0.6, colour = "steelblue") +
  geom_smooth(method = "loess", span = 0.6, se = FALSE,
              colour = "firebrick", linewidth = 0.9, formula = y ~ x) +
  labs(x = expression(z[1] == log(1 + "days since last insp."/365)),
       y = expression(widehat(TE)[i])) +
  theme_bw(base_size = 11)
```

@fig-cwa-eff-by-insp visualises the inspection-gap channel non-parametrically. The nearly flat, faintly upward LOESS slope mirrors the small and statistically weak negative $\hat{\delta}_1$ in @tbl-cwa-delta-marginals: in this sample the time since the last inspection does not move facilities measurably away from the feasible compliance minimum, contrary to the attention-lapse prior of @sec-emp-spec.

The empirical findings echo the simulation evidence of @sec-sim-scaling under the cost orientation. The data prefer a Gamma inefficiency distribution with scaling determinants over both the exponential restriction and the homogeneous baseline, consistent with the simulation finding that ignoring $\boldsymbol{\delta}$ when $\mathbf{z}_i$ matters produces upward bias in $\hat{b}$ and an over-compressed efficiency distribution. The marginal effects @eq-cwa-marginal are economically interpretable in non-compliance-quarter units, providing a quantitative basis for regulatory targeting: an additional unit of $z_{ij}$ shifts expected inefficiency proportionally, and the ranking of facilities by $\widehat{\mathrm{TE}}_i$ identifies the units that operate furthest above the feasible minimum given their observable characteristics.
