---
title: "Empirical application: patents and R&D"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Empirical application: patents and R&D}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "75%"
)
```

This vignette walks through the empirical application from
Centorrino and Perez Urdiales (2026), fitting a Poisson stochastic
frontier model to the canonical patents-R&D panel of Hausman, Hall,
and Griliches (1984) using `countSFA`.

## Data

The cleaned panel ships with the package:

```{r data}
library(countSFA)
data(patents)
str(patents)
```

There are 346 manufacturing firms observed over five years (1968-1972),
yielding 1730 firm-year observations. The outcome is the integer count
of patent applications eventually granted in each year.

```{r y-summary}
summary(patents$patents)
mean(patents$patents == 0)         # share of zero-patent observations
var(patents$patents) / mean(patents$patents)   # variance-to-mean ratio
```

The variance-to-mean ratio far exceeds 1, so a plain Poisson model
will under-state dispersion. The Poisson stochastic frontier
attributes the excess dispersion to one-sided technical inefficiency.

## Build the design matrix

We include log R&D, log capital, the science-sector indicator, and
year fixed effects:

```{r design}
X <- model.matrix(
  ~ log_rd + log_k + science_sector + factor(year),
  data = patents
)
y <- patents$patents
dim(X)
```

## Fit the models

### Pooled Poisson GLM (misspecified benchmark)

```{r poisson-glm}
fit_pois <- glm(y ~ X - 1, family = poisson())
ll_pois  <- as.numeric(logLik(fit_pois))
ll_pois
```

### Exponential frontier

```{r exp-frontier}
fit_exp <- fit_poisson_frontier(y, X, dist = "exponential")
fit_exp$loglik
fit_exp$b          # rate of the exponential inefficiency
fit_exp$K          # series truncation auto-selected from the GLM warm-start
```

The auto-K rule chose `r fit_exp$K` because the maximum fitted Poisson
mean drives the convergence boundary of the alternating series
(see `?fit_poisson_frontier`).

### Gamma frontier

The Gamma frontier on this dataset converges to alpha ~ 1, i.e.\ the
exponential boundary --- the data carry little information about the
shape parameter beyond what the exponential captures. This is the
weak-alpha-identification phenomenon documented in the paper's Monte
Carlo study, and on this particular sample the optimiser at the
boundary produces a singular Hessian, so we wrap the call in
`tryCatch`:

```{r gamma-frontier}
fit_gam <- tryCatch(
  suppressWarnings(fit_poisson_frontier(y, X, dist = "gamma")),
  error = function(e) {
    message("Gamma fit failed at the alpha boundary: ", conditionMessage(e))
    NULL
  }
)
if (!is.null(fit_gam)) {
  c(alpha = fit_gam$alpha, b = fit_gam$b, loglik = fit_gam$loglik)
}
```

## Compare the models

`compare_models()` returns a kable-ready summary:

```{r compare}
fits <- list(Exponential = fit_exp)
if (!is.null(fit_gam)) fits$Gamma <- fit_gam
compare_models(fits)
```

For non-nested comparisons (e.g.\ exponential frontier versus the
Poisson half-normal benchmark), use the Vuong (1989) test. The
half-normal model is fit by maximum simulated likelihood; the chunk
below is shown for reference but not executed at vignette build time
because the MSL fit on a real-world panel takes several minutes:

```{r halfnormal-vuong, eval = FALSE}
fit_phn <- fit_poisson_halfnormal(y, X)
v <- vuong_test(fit_exp, fit_phn, y, X)
v$statistic
v$p_value
v$decision
```

On the full patent panel, the exponential frontier is preferred over
the half-normal benchmark (V = 3.76, p = 0.0002).

## Posterior efficiency scores

For the exponential frontier, the posterior mean
\eqn{\mathrm{E}[e^{-u_i} \mid y_i, x_i]} has a closed form:

```{r efficiency}
scores <- efficiency_scores(fit_exp, y, X)
head(scores)
mean(scores$eff_score)
```

A simple histogram of the firm-level efficiency scores:

```{r eff-hist}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  ggplot(scores, aes(x = eff_score)) +
    geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
    geom_vline(xintercept = mean(scores$eff_score),
               linetype = "dashed", colour = "firebrick") +
    labs(x = "Technical efficiency score", y = "Count") +
    theme_bw()
}
```

The mass below 1 indicates that the average firm produces only a
fraction of its frontier patent output --- the central object of
interest in stochastic frontier analysis.

## References

- Centorrino, S. and Perez Urdiales, M. (2026). *Count Data Stochastic
  Frontier Models with Gamma Inefficiency*. Working paper.
- Fé, E. and Hofler, R. A. (2013). Count data stochastic frontier
  models, with an application to the patents-R&D relationship.
  *Journal of Productivity Analysis* **39**, 271-284.
- Hausman, J. A., Hall, B. H., and Griliches, Z. (1984). Econometric
  models for count data with an application to the patents-R&D
  relationship. *Econometrica* **52**, 909-938.
- Vuong, Q. H. (1989). Likelihood ratio tests for model selection and
  non-nested hypotheses. *Econometrica* **57**, 307-333.
